---
title: "Untitled"
format: html
editor: visual
---

> **Archived build (July 2026).** Every code chunk above the "Post-QA" section is commented out and set to `eval: false`. It is kept to show how the dictionaries were first built. Do not rerun it: it assigns new random `ref_id`s and would undo the October 2026 corrections. The current workflow is the "Post-QA" section at the end of this file.

# Set up

```{r}
#| eval: false
# library(tidyverse)
# library(readxl)
# library(stringr)
# library(stringi)
# library(ggraph)
# library(tidygraph)
# library(purrr)
# library(bib2df)
```

# Batch 1

## Batch 1a: Original 10 reviews

10 original reviews,

```{r}
#| eval: false
# batch1a <- read_excel("~/Documents/ODSi/ODSiData/citation-network/Doi_References/batch1a_dois_all.xlsx")%>%
#     transform(local_number = as.numeric(local_number))

```

## Batch 1b: Best selected (22 pubmed style)

22 selected best reviews with pubmed format

```{r}
#| eval: false
# batch1b<- read_excel("~/Documents/ODSi/ODSiData/citation-network/Doi_References/batch1b_dois_all.xlsx")%>%
#   select(-"doi_source")


# batch1 <- rbind(batch1a, batch1b)%>%
#     transform(local_number = as.numeric(local_number))
# table(table(batch1$doi))

# #rm(batch1a, batch1b)
```

# Batch 2: Pubmeds & non Pubmeds without abbreviations

## Batch 2a: Pubmed with rough abbreviations

```{r}
#| eval: false

# batch2a_path <- "~/Documents/ODSi/ODSiData/citation-network/Hand_Copied_References/batch2a_raw.xlsx"
# batch2a <- batch2a_path |>
#   excel_sheets() |>
#   set_names() |>
#   map(\(sheet) read_excel(batch2a_path, sheet = sheet)) |>
#   list_rbind(names_to = "sheet_name") %>%
#   group_by(sheet_name) %>%
#   fill(citing_paper, citing_doi, .direction = "down") %>%
#   mutate(local_number = row_number()) %>%
#   ungroup()%>%
#     transform(local_number = as.numeric(local_number))
```

## Batch 2b: Pubmeds without abbreviations1

```{r}
#| eval: false
# batch2b_path <- "~/Documents/ODSi/ODSiData/citation-network/Hand_Copied_References/batch2b_raw.xlsx"

# batch2b <- batch2b_path |>
#   excel_sheets() |>
#   set_names() |>
#   map(\(sheet) read_excel(batch2b_path, sheet = sheet)) |>
#   list_rbind(names_to = "sheet_name") %>%
#   group_by(sheet_name) %>%
#   fill(citing_paper, citing_doi, .direction = "down") %>%
#   mutate(local_number = row_number()) %>%
#   ungroup()%>%
#     transform(local_number = as.numeric(local_number))
```

## Batch 2c: NonPubmeds

```{r}
#| eval: false
# batch2c_path <- "~/Documents/ODSi/ODSiData/citation-network/Hand_Copied_References/batch2c_fixed.xlsx"
# batch2c <-batch2c_path |>
#   excel_sheets() |>
#   set_names() |>
#   map(\(sheet) read_excel(batch2c_path, sheet = sheet)) |>
#   list_rbind(names_to = "sheet_name") %>%
#   group_by(sheet_name) %>%
#   fill(citing_paper, citing_doi, .direction = "down") %>%

#   mutate(local_number = row_number()) %>%
#   ungroup()
```

## Batch 2 organizing

```{r}
#| eval: false
# batch2_combined <- bind_rows(batch2a, batch2b, batch2c)

# #write_csv(batch2_combined, file = "~/Documents/ODSi/ODSiData/citation-network/Organized_References.xlsx")

# #rm(batch2a, batch2b, batch2c, batch2_combined)
```

## Importing dois

```{r}
#| eval: false
# batch2 <- read_excel("~/Documents/ODSi/ODSiData/citation-network/Doi_References/batch2_dois_all.xlsx")%>%
#     transform(local_number = as.numeric(local_number))

```

# Batch 3: Correcting individual issues

```{r}
#| eval: false
# batch3_path <- "~/Documents/ODSi/ODSiData/citation-network/Hand_Copied_References/batch3_raw.xlsx"

# batch3 <- batch3_path |>
#   excel_sheets() |>
#   set_names() |>
#   map(\(sheet) read_excel(batch3_path, sheet = sheet)) |>
#   list_rbind(names_to = "sheet_name") %>%
#   group_by(sheet_name) %>%
#   fill(citing_paper, citing_doi, .direction = "down") %>%
#   mutate(local_number = row_number()) %>%
#   ungroup()%>%
#     transform(local_number = as.numeric(local_number))
```

# Catching Tricky DOI

Total 10,841 references

```{r}
#| eval: false
# refs_stage0 <- bind_rows(batch1, batch2, batch3)%>%
#   select(citing_paper, citing_doi, local_number, full_citation, doi)%>%
#     transform(local_number = as.numeric(local_number))

# #10841

# refs_fixed <- bind_rows(batch1, batch2, batch3)%>%
#   mutate(ID = paste0(citing_doi, "_", local_number))%>%
#   select(ID, citing_paper, citing_doi, local_number, full_citation, doi)%>%
#     transform(local_number = as.numeric(local_number))%>%
#   filter(!(citing_paper %in% "Zhou&Fang_2026"))%>%
#   distinct(ID, .keep_all = TRUE)
# #10823
```

## Stage 1

```{r}
#| eval: false
# refs_stage1 <- refs_stage0 %>%
#     filter(!is.na(doi))

# #9168
  
# needsDOI_stage1<- refs_stage0 %>% filter(is.na(doi))
# #1673

# #Total 10841
  
#  # write_csv(needsDOI_stage1, file = "~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/missed_dois1.csv")


```

## Stage 2

```{r}
#| eval: false
# import_stage2 <- read_excel("~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/reran_dois1.xlsx")%>%
#     transform(local_number = as.numeric(local_number))

# refs_stage2 <- bind_rows(refs_stage1, import_stage2)%>%
#     filter(!is.na(doi)| grepl("US National Library of Medicine", full_citation))%>%
#   select(citing_paper, citing_doi, local_number, full_citation, doi)
# #9586



# needsDOI_stage2_all<- import_stage2 %>% 
#   filter(is.na(doi)&
#            !grepl("US National Library of Medicine", full_citation))%>%
#   select(citing_paper, citing_doi, local_number, full_citation, doi)%>%
#   group_by(citing_doi, citing_paper)%>%
#   mutate(N = n())
# #1255

# needsDOI_stage2_a <- needsDOI_stage2_all %>%
#   filter(N > 10)%>%
#   filter(citing_paper != "Shono&vandenBrink_2018")
# #1080

# needsDOI_stage2_b <- needsDOI_stage2_all %>%
#   filter(N <= 10)%>%
#   filter(citing_paper != "Shono&vandenBrink_2018")
# #172

# needsDOI_stage2_c <- needsDOI_stage2_all %>%
#   filter(citing_paper == "Shono&vandenBrink_2018")
# #3

# # Total 10841

  
#  # write_csv(needsDOI_stage2_ai, file = "~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/missed_dois2_a.csv")
  
#  #   write_csv(needsDOI_stage2_hand, file = "~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/missed_dois2_b.csv")
```

## Stage 3

```{r}
#| eval: false
# import_stage3 <- read_excel("~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/reran_dois2.xlsx")%>%
#   transform(local_number = as.numeric(local_number))

# refs_stage3 <- bind_rows(
#   refs_stage2 %>%
#     filter(!is.na(doi) & !grepl("US National Library of Medicine", full_citation))%>%
#     transform(local_number = as.numeric(local_number))%>%
#     select(citing_paper, citing_doi, local_number, full_citation, doi),
#   #9557
#   import_stage3%>%
#     filter(!is.na(doi) & doi_source != "no_doi_expected")%>%
#     select(citing_paper, citing_doi, local_number, full_citation, doi)
#   #1074
#   )
# #10631

# noDOI_stage3 <- bind_rows(
#   refs_stage2 %>%
#     filter(grepl("US National Library of Medicine", full_citation)|
#              is.na(doi))%>%
#     transform(local_number = as.numeric(local_number))%>%
#     select(citing_paper, citing_doi, local_number, full_citation, doi),
#   #29
#   import_stage3 %>%
#     filter(doi_source == "no_doi_expected")%>%
#     select(citing_paper, citing_doi, local_number, full_citation, doi)
#   )
# #65

# needsDOI_stage3<- import_stage3 %>% 
#   filter(is.na(doi)&
#            doi_source != "no_doi_expected")%>%
#   select(citing_paper, citing_doi, local_number, full_citation, doi)%>%
#   group_by(citing_doi, citing_paper)
# #145

# #Total 10841

# #write_csv(needsDOI_stage3, file = "~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/missed_dois3.csv")

```

## Stage 4

```{r}
#| eval: false
# import_stage4 <- read_excel("~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/reran_dois3.xlsx")%>%
#   transform(local_number = as.numeric(local_number))

# #Normal DOI
# refs_stage4 <- bind_rows(
#   refs_stage3 ,
#   import_stage4 %>%
#     filter(!is.na(doi))%>%
#   select(citing_paper, citing_doi, local_number, full_citation, doi)
#   )%>%
#   mutate(ID = paste0(citing_doi, "_", local_number))
# #10691

# #Alternative ID
# refs_altid <- import_stage4 %>%
#   filter(is.na(doi) &
#            is.na(Other) &
#           ( !(is.na(PMID)) | 
#               !is.na(PMCID) |
#               !is.na(ISBN) |
#               !is.na(ISSN) |
#               !is.na(CRID)
#           )
#   )%>%
#   mutate(ID = paste0(citing_doi, "_", local_number))
# #70

# refs_clinicaltrials <- rbind(noDOI_stage3 ,
#                   import_stage4 %>%
#                     filter(!is.na(Other)&
#                              !(Other %in% c("DNE",
#       "TYPO_https://www.ncbi.nlm.nih.gov/books/NBK538168/",
#       "TYPO_OR_DNE")))%>%
#         select(citing_paper, citing_doi, local_number, full_citation, doi)
#       )%>%
#   filter(grepl("US National Library", full_citation))%>%
#   mutate(NCT = str_extract(full_citation, "NCT\\d{8}"))%>%
#   select(citing_paper, citing_doi, local_number, full_citation, doi, NCT)%>%
#   mutate(ID = paste0(citing_doi, "_", local_number))
# #29

# #Websites or no other info
# refs_web <- rbind(noDOI_stage3 %>%
#                     mutate(Website = NA),
#                   import_stage4 %>%
#                     filter(!is.na(Other)&
#                              !(Other %in% c("DNE",
#       "TYPO_https://www.ncbi.nlm.nih.gov/books/NBK538168/",
#       "TYPO_OR_DNE")))%>%
#         mutate(Website = Other)%>%
#         select(citing_paper, citing_doi, local_number, full_citation, doi, Website)
#         )%>%
#   filter(!grepl("US National Library", full_citation))%>%
#   mutate(ID = paste0(citing_doi, "_", local_number))
# #48



# refs_anomaly <- import_stage4 %>%
#   filter(Other %in% c("DNE",
#       "TYPO_https://www.ncbi.nlm.nih.gov/books/NBK538168/",
#       "TYPO_OR_DNE"))%>%
#   mutate(ID = paste0(citing_doi, "_", local_number))
# #3

# #10841
```

## Stage 5: Adding in Wang&Zhou_2025 and Qayad&Horan and removing duplicated Chen_2015 and Zhou&Feng

```{r}
#| eval: false
# refs_stage5 <- rbind(
#   refs_stage4 %>%
#   distinct(ID, .keep_all = TRUE),
#  batch3 %>% filter(!is.na(doi) & 
#                      citing_paper %in% c("Wang&Zhou_2025",
#                                          "Qayed&Horan_2015"))%>%
#   mutate(ID = paste0(citing_doi, "_", local_number))%>%
#    select(ID, citing_paper, citing_doi, local_number, full_citation, doi)
# )
# #10691 -> 10595 -> 

# refs_altid <- refs_altid %>%
#   bind_rows(batch3 %>%
#               filter(is.na(doi)& 
#                      citing_paper %in% c("Wang&Zhou_2025",
#                                          "Qayed&Horan_2015"))%>%
#   mutate(ID = paste0(citing_doi, "_", local_number)))



```

# Identifying Together

```{r}
#| eval: false

# master <- refs_fixed %>%
#   mutate(ID = paste0(citing_doi, "_", local_number)) %>%
#   select(ID, citing_paper, local_number, full_citation)
# #10823

# #DOI
# all_identifiers <- bind_rows(
#   # First identifier source (DOIs)
#   refs_stage5 %>%
#     mutate(Type = "Paper",
#            TypeOfIdentifier = "DOI",
#            Identifier = as.character(doi)) %>%
#     select(ID, Type, TypeOfIdentifier, Identifier),
  
# #PMID
#   refs_altid %>%
#     filter(!is.na(PMID)) %>%
#     mutate(Type = "Paper",
#            TypeOfIdentifier = "PMID",
#            Identifier = as.character(PMID)) %>%
#     select(ID, Type, TypeOfIdentifier, Identifier),

# #ISBN
#   refs_altid %>%
#     filter(!is.na(ISBN)) %>%
#     mutate(Type = "Book",
#            TypeOfIdentifier = "ISBN",
#            Identifier = as.character(ISBN)) %>%
#     select(ID, Type, TypeOfIdentifier, Identifier),

# #CRID
#   refs_altid %>%
#     filter(!is.na(CRID)) %>%
#     mutate(Type = "ClinicalResearch",
#            TypeOfIdentifier = "CRID",
#            Identifier = as.character(ISBN)) %>%
#     select(ID, Type, TypeOfIdentifier, Identifier),

# #Clinical Trials
#   refs_clinicaltrials %>%
#     mutate(Type = "ClinicalResearch",
#            TypeOfIdentifier = "NCT",
#            Identifier = as.character(NCT)) %>%
#     select(ID, Type, TypeOfIdentifier, Identifier)
  
  
# ) %>% 
#   distinct() # Removes any exact duplicates across the combined tables

# # 3. Join the combined identifiers to master in one step
# master_itn <- master %>%
#   left_join(all_identifiers, by = "ID")

# # write_csv(master_itn %>%
# #             filter(is.na(Identifier)),
# #           file = "~/Documents/ODSi/ODSiData/citation-network/handcheck.csv")

# master_auto <- master_itn %>%
#   filter(!is.na(Identifier))
# #10772

# master_handcheck <- read_csv("~/Documents/ODSi/ODSiData/citation-network/handcheck.csv")%>%distinct(ID, .keep_all = TRUE)
# #51

# master_final <- bind_rows(master_auto, master_handcheck)
# #10823
```

# Finalizing Identifying Info

```{r}
#| eval: false

# bib <- bib2df::bib2df(file = "~/Documents/ODSi/ODSiData/citation-network/Full_Network/full_bib.bib")
```

# Foreign Key

```{r}
#| eval: false
# citing_dict <- master_final %>%
#   mutate(call_id = paste0(citing_paper, "_", local_number))%>%
#   left_join(refs_fixed %>%
#               select(citing_paper, citing_doi)%>%
#               distinct(),
#             by = "citing_paper")

# set.seed(9134)
# citing_dict$id = ids::random_id(n = nrow(citing_dict), bytes = 4)

# citing_dict <- citing_dict %>%
#   select(id, call_id, citing_paper, citing_doi, local_number, full_citation, Type, TypeOfIdentifier, Identifier)
```

```{r}
#| eval: false
# linking_dict <- bib %>%
#   select(BIBTEXKEY, DOI, PMID, ISSN, ISBN, TITLE, AUTHOR)

# linking_dict$ref_id <- ids::random_id(n = nrow(linking_dict), bytes = 4)
```

```{r}
#| eval: false
# citing_dict2 <- citing_dict %>%
#   left_join(linking_dict %>%
#               select(citing_bibtex = BIBTEXKEY, DOI, citing_id = ref_id),
#             by = c("citing_doi" = "DOI" ))%>%
#   select(id, call_id, citing_bibtex, citing_id, local_number, full_citation, Type, TypeOfIdentifier, Identifier)
```

```{r}
#| eval: false
# linking_long <- linking_dict %>%
#   # Select the ID column and the target identifier columns
#   select(ref_id, ref_bibtex = BIBTEXKEY, DOI, PMID, ISBN) %>%
#   # Pivot the data so column names become a variable and values become a variable
#   pivot_longer(
#     cols = c(DOI, PMID, ISBN),
#     names_to = "TypeOfIdentifier",
#     values_to = "Identifier",
#     # Ensure all identifiers are treated as characters to avoid type-mismatch errors
#     values_transform = list(Identifier = as.character)
#   ) %>%
#   # Drop empty rows where no identifier existed
#   filter(!is.na(Identifier) & Identifier != "") %>%
#   # Standardize case: TypeOfIdentifier to uppercase, Identifier to lowercase
#   mutate(
#     TypeOfIdentifier = toupper(TypeOfIdentifier),
#     Identifier = tolower(Identifier)  # <--- FIX ADDED HERE
#   )

# # 2. Join the reshaped dictionary to your citing dictionary
# citing_dict3_linked <- citing_dict2 %>%
#   # Standardize the citing_dict2 Identifier to lowercase as well
#   mutate(
#     Identifier = tolower(as.character(Identifier)) # <--- FIX ADDED HERE
#   ) %>%
#   # Perform a left join to bring in ref_id where there is a match
#   left_join(linking_long, by = c("TypeOfIdentifier", "Identifier"))

# unlinked <- citing_dict3_linked %>%
#   filter(is.na(ref_id))
```

# Export

-   Zotero library (.bib)

-   citing_dictionary

-   ref_dictionary

```{r}
#| eval: false
# citing_dictionary <- citing_dict3_linked
# # save(citing_dictionary, file = "~/Documents/ODSi/ODSiData/citation-network/Full_Network/citing_dictionary.Rdata")
# # write_excel_csv(citing_dictionary, file = "~/Documents/ODSi/ODSiData/citation-network/Full_Network/citing_dictionary.csv" )
```

```{r}
#| eval: false
# # ref_dictionary <- linking_dict %>%
# #   select(ref_id, names(linking_dict)[1:7])
# #  save(ref_dictionary, file = "~/Documents/ODSi/ODSiData/citation-network/Full_Network/ref_dictionary.Rdata")
# #  write_excel_csv(ref_dictionary, file = "~/Documents/ODSi/ODSiData/citation-network/Full_Network/ref_dictionary.csv" )
```

# Post-QA (October 2026): corrected dictionaries and BibTeX export

Everything above this section is the original July 2026 build, kept for the record and commented out so it cannot be rerun by accident. Rerunning it would assign new random `ref_id`s and undo the October 2026 corrections.

Starting point: the dictionaries as left by `parsing_scripts/apply_dictionary_corrections.R` (steps 1 to 3, applied 2026-10-06; originals in `Full_Network/backup_20261006_171459/`). This section

1.  restores `YEAR` and `MONTH`, which `bib2df` dropped on import because Zotero writes them unbraced (`year = 2001`);
2.  fills thin entries (no author, journal or year) that are cited and have a DOI, from Crossref, filling blank fields only;
3.  syncs `ref_dictionary` with `bib`, runs checks, and writes `citing_dictionary`, `ref_dictionary` and `bib` (`.Rdata` + `.csv`) plus `full_bib_corrected.bib`.

Safe to rerun: Crossref answers are cached in `corrections/crossref_fill_cache.rds`, and only blank fields are ever filled. It does not use `bib2df`.

```{r post-qa-setup}
library(tidyverse)
library(jsonlite)

fn   <- "~/Documents/ODSi/ODSiData/citation-network/Full_Network"
corr <- file.path(fn, "corrections")

load(file.path(fn, "citing_dictionary.Rdata"))
load(file.path(fn, "ref_dictionary.Rdata"))
load(file.path(fn, "bib.Rdata"))
stopifnot(nrow(ref_dictionary) == nrow(bib),
          identical(ref_dictionary$BIBTEXKEY, bib$BIBTEXKEY))

blank <- function(x) is.na(x) | x == "" | x == "NA"
no_author <- function(a) vapply(a, function(v) all(is.na(v)) || all(v == ""), logical(1))

# backup of the inputs to this section
stamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
bk <- file.path(fn, paste0("backup_", stamp, "_postqa"))
dir.create(bk)
for (f in c("citing_dictionary", "ref_dictionary", "bib"))
  for (x in c(".Rdata", ".csv")) file.copy(file.path(fn, paste0(f, x)), bk)
```

## Restore YEAR and MONTH from the Zotero export

`bib` rows 1 to 5,680 are the entries of `full_bib.bib` in file order (later rows were added during the corrections). Each value is copied only when the row's key or title still matches the file entry at that position.

```{r post-qa-year}
txt <- paste(readLines(file.path(fn, "full_bib.bib"), encoding = "UTF-8", warn = FALSE),
             collapse = "\n")
ent <- strsplit(txt, "\n(?=@)", perl = TRUE)[[1]]
ent <- ent[grepl("^@", ent)]
bib_field <- function(f) {
  m <- regmatches(ent, regexec(paste0("\\n\\s*", f, " = \\{?([^,}\\n]*)\\}?,?\\n"), ent))
  vapply(m, function(x) if (length(x)) x[2] else NA_character_, "")
}
file_key   <- sub("^@[A-Za-z]+\\{([^,]*),.*", "\\1", ent)
file_title <- sub(".*\\n\\s*title = \\{(.*?)\\},?\\n.*", "\\1", ent)
file_year  <- bib_field("year")
file_month <- bib_field("month")

n_file <- length(ent)
norm <- function(x) tolower(gsub("[^A-Za-z0-9]", "", x))
same <- file_key == bib$BIBTEXKEY[seq_len(n_file)] |
        norm(file_title) == norm(bib$TITLE[seq_len(n_file)])
cat("file entries:", n_file, "| aligned:", sum(same), "\n")
stopifnot(mean(same) > 0.99)

i <- which(same & blank(bib$YEAR[seq_len(n_file)]) & !is.na(file_year))
bib$YEAR[i] <- file_year[i]
j <- which(same & blank(bib$MONTH[seq_len(n_file)]) & !is.na(file_month))
bib$MONTH[j] <- file_month[j]
cat("YEAR restored:", length(i), "| MONTH restored:", length(j), "\n")
```

## Crossref fill for thin entries

Thin = cited at least once, has a real DOI (not a Faculty Opinions `10.3410/` record), and lacks an author, journal or year. Only blank fields are filled. Every filled value is logged to `corrections/crossref_fill_log.csv`.

```{r post-qa-crossref}
cache_path <- file.path(corr, "crossref_fill_cache.rds")
cache <- if (file.exists(cache_path)) readRDS(cache_path) else list()

cr_get <- function(doi) {
  if (!is.null(cache[[doi]])) return(cache[[doi]])
  url <- paste0("https://api.crossref.org/works/", utils::URLencode(doi, reserved = TRUE),
                "?mailto=sophhuebler@gmail.com")
  msg <- tryCatch(jsonlite::fromJSON(url, simplifyVector = FALSE)$message,
                  error = function(e) NULL)
  Sys.sleep(0.5)                       # polite pool
  if (!is.null(msg)) cache[[doi]] <<- msg
  msg
}
`%||%` <- function(a, b) if (is.null(a) || !length(a)) b else a
cr_fields <- function(m) {
  au <- vapply(m$author %||% list(), function(a)
    if (!is.null(a$family)) trimws(paste0(a$family, ", ", a$given %||% "")) else a$name %||% "", "")
  list(AUTHOR  = if (length(au)) au else NULL,
       YEAR    = as.character(m$issued$`date-parts`[[1]][[1]] %||% NA),
       JOURNAL = unlist(m$`container-title`)[1] %||% NA,
       VOLUME  = m$volume %||% NA,
       NUMBER  = m$issue %||% NA,
       PAGES   = gsub("-", "--", m$page %||% NA),
       TITLE   = unlist(m$title)[1] %||% NA)
}

cited <- bib$BIBTEXKEY %in% citing_dictionary$ref_bibtex
real_doi <- !blank(bib$DOI) & !grepl("^10\\.3410/", bib$DOI)
thin <- which(cited & real_doi &
              (no_author(bib$AUTHOR) | blank(bib$JOURNAL) | blank(bib$YEAR)))
cat("thin cited entries with a DOI:", length(thin), "\n")

fill_log <- list()
for (k in thin) {
  m <- cr_get(bib$DOI[k])
  if (is.null(m)) { fill_log[[length(fill_log) + 1]] <- tibble(BIBTEXKEY = bib$BIBTEXKEY[k],
      field = "(lookup failed)", old = NA, new = bib$DOI[k]); next }
  v <- cr_fields(m)
  container_col <- if (bib$CATEGORY[k] %in% c("INCOLLECTION", "INBOOK")) "BOOKTITLE" else "JOURNAL"
  for (f in names(v)) {
    col <- if (f == "JOURNAL") container_col else f
    new <- v[[f]]
    if (is.null(new) || all(is.na(new))) next
    if (col == "AUTHOR") {
      if (!no_author(bib$AUTHOR[k])) next
      bib$AUTHOR[[k]] <- new
      new <- paste(new, collapse = " and ")
    } else {
      if (!blank(bib[[col]][k])) next
      bib[[col]][k] <- new
    }
    fill_log[[length(fill_log) + 1]] <- tibble(BIBTEXKEY = bib$BIBTEXKEY[k], field = col,
                                               old = NA, new = as.character(new))
  }
}
saveRDS(cache, cache_path)
fill_log <- bind_rows(fill_log)
write_csv(fill_log, file.path(corr, paste0("crossref_fill_log_", stamp, ".csv")))
count(fill_log, field)
```

## Sync, check and write

```{r post-qa-sync}
ref_dictionary$TITLE  <- bib$TITLE
ref_dictionary$DOI    <- bib$DOI
ref_dictionary$PMID   <- bib$PMID
ref_dictionary$AUTHOR <- bib$AUTHOR

chk <- tibble(
  check = c("ref_dictionary and bib aligned", "unique ref_id", "unique BIBTEXKEY",
            "citing rows whose ref_id is missing from ref_dictionary",
            "citing rows linked to Faculty Opinions records",
            "citing rows with NA citing_bibtex",
            "cited entries still missing year", "cited entries still missing author"),
  value = c(identical(ref_dictionary$BIBTEXKEY, bib$BIBTEXKEY),
            !anyDuplicated(ref_dictionary$ref_id), !anyDuplicated(bib$BIBTEXKEY),
            sum(!blank(citing_dictionary$ref_id) & !citing_dictionary$ref_id %in% ref_dictionary$ref_id),
            sum(grepl("^10\\.3410/", ref_dictionary$DOI[match(citing_dictionary$ref_id, ref_dictionary$ref_id)])),
            sum(is.na(citing_dictionary$citing_bibtex)),
            sum(cited & blank(bib$YEAR)), sum(cited & no_author(bib$AUTHOR))))
chk
stopifnot(chk$value[1:3] == 1, chk$value[4:6] == 0)
```

```{r post-qa-write}
flat <- function(df) mutate(df, across(where(is.list), ~ vapply(.x, function(v)
  if (all(is.na(v))) "" else paste(v, collapse = " and "), "")))

# BibTeX writer (bib2df mis-reads Zotero's unbraced year/month, so it is not used)
write_bib <- function(b, ids, path) {
  skip <- c("CATEGORY", "BIBTEXKEY")
  one <- function(k) {
    vals <- c(lapply(setdiff(names(b), skip), function(f) {
      v <- b[[f]][[k]]
      if (is.list(v)) v <- unlist(v)
      if (all(is.na(v)) || all(v == "") || all(v == "NA")) return(NULL)
      v <- if (f %in% c("AUTHOR", "EDITOR")) paste(v, collapse = " and ") else paste(v, collapse = ", ")
      if (f == "MONTH" && grepl("^[a-z]{3}$", v)) return(paste0("  month = ", v))  # BibTeX month macro
      paste0("  ", tolower(f), " = {", v, "}")
    }), list(paste0("  refid = {", ids[k], "}")))
    paste0("@", tolower(b$CATEGORY[k]), "{", b$BIBTEXKEY[k], ",\n",
           paste(unlist(vals), collapse = ",\n"), "\n}\n")
  }
  writeLines(vapply(seq_len(nrow(b)), one, ""), path, useBytes = TRUE)
}


```

```{r}
save(citing_dictionary, file = file.path(fn, "citing_dictionary.Rdata"))
save(ref_dictionary,    file = file.path(fn, "ref_dictionary.Rdata"))
save(bib,               file = file.path(fn, "bib.Rdata"))
write_excel_csv(flat(citing_dictionary), file.path(fn, "citing_dictionary.csv"))
write_excel_csv(flat(ref_dictionary),    file.path(fn, "ref_dictionary.csv"))
write_excel_csv(flat(bib),               file.path(fn, "bib.csv"))
write_bib(bib, ref_dictionary$ref_id, file.path(fn, "full_bib_corrected.bib"))
cat("Wrote dictionaries (.Rdata, .csv) and full_bib_corrected.bib; inputs backed up to", bk, "\n")
```
