---
title: "Untitled"
format: html
editor: visual
---

# Set up

```{r}
library(tidyverse)
library(readxl)
library(stringr)
library(stringi)
library(ggraph)
library(tidygraph)
library(purrr)
library(bib2df)
```

# Batch 1

## Batch 1a: Original 10 reviews

10 original reviews,

```{r}
batch1a <- read_excel("~/Documents/ODSi/ODSiData/citation-network/Doi_References/batch1a_dois_all.xlsx")%>%
    transform(local_number = as.numeric(local_number))

```

## Batch 1b: Best selected (22 pubmed style)

22 selected best reviews with pubmed format

```{r}
batch1b<- read_excel("~/Documents/ODSi/ODSiData/citation-network/Doi_References/batch1b_dois_all.xlsx")%>%
  select(-"doi_source")


batch1 <- rbind(batch1a, batch1b)%>%
    transform(local_number = as.numeric(local_number))
table(table(batch1$doi))

#rm(batch1a, batch1b)
```

# Batch 2: Pubmeds & non Pubmeds without abbreviations

## Batch 2a: Pubmed with rough abbreviations

```{r}

batch2a_path <- "~/Documents/ODSi/ODSiData/citation-network/Hand_Copied_References/batch2a_raw.xlsx"
batch2a <- batch2a_path |>
  excel_sheets() |>
  set_names() |>
  map(\(sheet) read_excel(batch2a_path, sheet = sheet)) |>
  list_rbind(names_to = "sheet_name") %>%
  group_by(sheet_name) %>%
  fill(citing_paper, citing_doi, .direction = "down") %>%
  mutate(local_number = row_number()) %>%
  ungroup()%>%
    transform(local_number = as.numeric(local_number))
```

## Batch 2b: Pubmeds without abbreviations1

```{r}
batch2b_path <- "~/Documents/ODSi/ODSiData/citation-network/Hand_Copied_References/batch2b_raw.xlsx"

batch2b <- batch2b_path |>
  excel_sheets() |>
  set_names() |>
  map(\(sheet) read_excel(batch2b_path, sheet = sheet)) |>
  list_rbind(names_to = "sheet_name") %>%
  group_by(sheet_name) %>%
  fill(citing_paper, citing_doi, .direction = "down") %>%
  mutate(local_number = row_number()) %>%
  ungroup()%>%
    transform(local_number = as.numeric(local_number))
```

## Batch 2c: NonPubmeds

```{r}
batch2c_path <- "~/Documents/ODSi/ODSiData/citation-network/Hand_Copied_References/batch2c_fixed.xlsx"
batch2c <-batch2c_path |>
  excel_sheets() |>
  set_names() |>
  map(\(sheet) read_excel(batch2c_path, sheet = sheet)) |>
  list_rbind(names_to = "sheet_name") %>%
  group_by(sheet_name) %>%
  fill(citing_paper, citing_doi, .direction = "down") %>%

  mutate(local_number = row_number()) %>%
  ungroup()
```

## Batch 2 organizing

```{r}
batch2_combined <- bind_rows(batch2a, batch2b, batch2c)

#write_csv(batch2_combined, file = "~/Documents/ODSi/ODSiData/citation-network/Organized_References.xlsx")

#rm(batch2a, batch2b, batch2c, batch2_combined)
```

## Importing dois

```{r}
batch2 <- read_excel("~/Documents/ODSi/ODSiData/citation-network/Doi_References/batch2_dois_all.xlsx")%>%
    transform(local_number = as.numeric(local_number))

```

# Batch 3: Correcting individual issues

```{r}
batch3_path <- "~/Documents/ODSi/ODSiData/citation-network/Hand_Copied_References/batch3_raw.xlsx"

batch3 <- batch3_path |>
  excel_sheets() |>
  set_names() |>
  map(\(sheet) read_excel(batch3_path, sheet = sheet)) |>
  list_rbind(names_to = "sheet_name") %>%
  group_by(sheet_name) %>%
  fill(citing_paper, citing_doi, .direction = "down") %>%
  mutate(local_number = row_number()) %>%
  ungroup()%>%
    transform(local_number = as.numeric(local_number))
```

# Catching Tricky DOI

Total 10,841 references

```{r}
refs_stage0 <- bind_rows(batch1, batch2, batch3)%>%
  select(citing_paper, citing_doi, local_number, full_citation, doi)%>%
    transform(local_number = as.numeric(local_number))

#10841

refs_fixed <- bind_rows(batch1, batch2, batch3)%>%
  mutate(ID = paste0(citing_doi, "_", local_number))%>%
  select(ID, citing_paper, citing_doi, local_number, full_citation, doi)%>%
    transform(local_number = as.numeric(local_number))%>%
  filter(!(citing_paper %in% "Zhou&Fang_2026"))%>%
  distinct(ID, .keep_all = TRUE)
#10823
```

## Stage 1

```{r}
refs_stage1 <- refs_stage0 %>%
    filter(!is.na(doi))

#9168
  
needsDOI_stage1<- refs_stage0 %>% filter(is.na(doi))
#1673

#Total 10841
  
 # write_csv(needsDOI_stage1, file = "~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/missed_dois1.csv")


```

## Stage 2

```{r}
import_stage2 <- read_excel("~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/reran_dois1.xlsx")%>%
    transform(local_number = as.numeric(local_number))

refs_stage2 <- bind_rows(refs_stage1, import_stage2)%>%
    filter(!is.na(doi)| grepl("US National Library of Medicine", full_citation))%>%
  select(citing_paper, citing_doi, local_number, full_citation, doi)
#9586



needsDOI_stage2_all<- import_stage2 %>% 
  filter(is.na(doi)&
           !grepl("US National Library of Medicine", full_citation))%>%
  select(citing_paper, citing_doi, local_number, full_citation, doi)%>%
  group_by(citing_doi, citing_paper)%>%
  mutate(N = n())
#1255

needsDOI_stage2_a <- needsDOI_stage2_all %>%
  filter(N > 10)%>%
  filter(citing_paper != "Shono&vandenBrink_2018")
#1080

needsDOI_stage2_b <- needsDOI_stage2_all %>%
  filter(N <= 10)%>%
  filter(citing_paper != "Shono&vandenBrink_2018")
#172

needsDOI_stage2_c <- needsDOI_stage2_all %>%
  filter(citing_paper == "Shono&vandenBrink_2018")
#3

# Total 10841

  
 # write_csv(needsDOI_stage2_ai, file = "~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/missed_dois2_a.csv")
  
 #   write_csv(needsDOI_stage2_hand, file = "~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/missed_dois2_b.csv")
```

## Stage 3

```{r}
import_stage3 <- read_excel("~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/reran_dois2.xlsx")%>%
  transform(local_number = as.numeric(local_number))

refs_stage3 <- bind_rows(
  refs_stage2 %>%
    filter(!is.na(doi) & !grepl("US National Library of Medicine", full_citation))%>%
    transform(local_number = as.numeric(local_number))%>%
    select(citing_paper, citing_doi, local_number, full_citation, doi),
  #9557
  import_stage3%>%
    filter(!is.na(doi) & doi_source != "no_doi_expected")%>%
    select(citing_paper, citing_doi, local_number, full_citation, doi)
  #1074
  )
#10631

noDOI_stage3 <- bind_rows(
  refs_stage2 %>%
    filter(grepl("US National Library of Medicine", full_citation)|
             is.na(doi))%>%
    transform(local_number = as.numeric(local_number))%>%
    select(citing_paper, citing_doi, local_number, full_citation, doi),
  #29
  import_stage3 %>%
    filter(doi_source == "no_doi_expected")%>%
    select(citing_paper, citing_doi, local_number, full_citation, doi)
  )
#65

needsDOI_stage3<- import_stage3 %>% 
  filter(is.na(doi)&
           doi_source != "no_doi_expected")%>%
  select(citing_paper, citing_doi, local_number, full_citation, doi)%>%
  group_by(citing_doi, citing_paper)
#145

#Total 10841

#write_csv(needsDOI_stage3, file = "~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/missed_dois3.csv")

```

## Stage 4

```{r}
import_stage4 <- read_excel("~/Documents/ODSi/ODSiData/citation-network/Catch_Doi/reran_dois3.xlsx")%>%
  transform(local_number = as.numeric(local_number))

#Normal DOI
refs_stage4 <- bind_rows(
  refs_stage3 ,
  import_stage4 %>%
    filter(!is.na(doi))%>%
  select(citing_paper, citing_doi, local_number, full_citation, doi)
  )%>%
  mutate(ID = paste0(citing_doi, "_", local_number))
#10691

#Alternative ID
refs_altid <- import_stage4 %>%
  filter(is.na(doi) &
           is.na(Other) &
          ( !(is.na(PMID)) | 
              !is.na(PMCID) |
              !is.na(ISBN) |
              !is.na(ISSN) |
              !is.na(CRID)
          )
  )%>%
  mutate(ID = paste0(citing_doi, "_", local_number))
#70

refs_clinicaltrials <- rbind(noDOI_stage3 ,
                  import_stage4 %>%
                    filter(!is.na(Other)&
                             !(Other %in% c("DNE",
      "TYPO_https://www.ncbi.nlm.nih.gov/books/NBK538168/",
      "TYPO_OR_DNE")))%>%
        select(citing_paper, citing_doi, local_number, full_citation, doi)
      )%>%
  filter(grepl("US National Library", full_citation))%>%
  mutate(NCT = str_extract(full_citation, "NCT\\d{8}"))%>%
  select(citing_paper, citing_doi, local_number, full_citation, doi, NCT)%>%
  mutate(ID = paste0(citing_doi, "_", local_number))
#29

#Websites or no other info
refs_web <- rbind(noDOI_stage3 %>%
                    mutate(Website = NA),
                  import_stage4 %>%
                    filter(!is.na(Other)&
                             !(Other %in% c("DNE",
      "TYPO_https://www.ncbi.nlm.nih.gov/books/NBK538168/",
      "TYPO_OR_DNE")))%>%
        mutate(Website = Other)%>%
        select(citing_paper, citing_doi, local_number, full_citation, doi, Website)
        )%>%
  filter(!grepl("US National Library", full_citation))%>%
  mutate(ID = paste0(citing_doi, "_", local_number))
#48



refs_anomaly <- import_stage4 %>%
  filter(Other %in% c("DNE",
      "TYPO_https://www.ncbi.nlm.nih.gov/books/NBK538168/",
      "TYPO_OR_DNE"))%>%
  mutate(ID = paste0(citing_doi, "_", local_number))
#3

#10841
```

## Stage 5: Adding in Wang&Zhou_2025 and Qayad&Horan and removing duplicated Chen_2015 and Zhou&Feng

```{r}
refs_stage5 <- rbind(
  refs_stage4 %>%
  distinct(ID, .keep_all = TRUE),
 batch3 %>% filter(!is.na(doi) & 
                     citing_paper %in% c("Wang&Zhou_2025",
                                         "Qayed&Horan_2015"))%>%
  mutate(ID = paste0(citing_doi, "_", local_number))%>%
   select(ID, citing_paper, citing_doi, local_number, full_citation, doi)
)
#10691 -> 10595 -> 

refs_altid <- refs_altid %>%
  bind_rows(batch3 %>%
              filter(is.na(doi)& 
                     citing_paper %in% c("Wang&Zhou_2025",
                                         "Qayed&Horan_2015"))%>%
  mutate(ID = paste0(citing_doi, "_", local_number)))



```

# Identifying Together

```{r}

master <- refs_fixed %>%
  mutate(ID = paste0(citing_doi, "_", local_number)) %>%
  select(ID, citing_paper, local_number, full_citation)
#10823

#DOI
all_identifiers <- bind_rows(
  # First identifier source (DOIs)
  refs_stage5 %>%
    mutate(Type = "Paper",
           TypeOfIdentifier = "DOI",
           Identifier = as.character(doi)) %>%
    select(ID, Type, TypeOfIdentifier, Identifier),
  
#PMID
  refs_altid %>%
    filter(!is.na(PMID)) %>%
    mutate(Type = "Paper",
           TypeOfIdentifier = "PMID",
           Identifier = as.character(PMID)) %>%
    select(ID, Type, TypeOfIdentifier, Identifier),

#ISBN
  refs_altid %>%
    filter(!is.na(ISBN)) %>%
    mutate(Type = "Book",
           TypeOfIdentifier = "ISBN",
           Identifier = as.character(ISBN)) %>%
    select(ID, Type, TypeOfIdentifier, Identifier),

#CRID
  refs_altid %>%
    filter(!is.na(CRID)) %>%
    mutate(Type = "ClinicalResearch",
           TypeOfIdentifier = "CRID",
           Identifier = as.character(ISBN)) %>%
    select(ID, Type, TypeOfIdentifier, Identifier),

#Clinical Trials
  refs_clinicaltrials %>%
    mutate(Type = "ClinicalResearch",
           TypeOfIdentifier = "NCT",
           Identifier = as.character(NCT)) %>%
    select(ID, Type, TypeOfIdentifier, Identifier)
  
  
) %>% 
  distinct() # Removes any exact duplicates across the combined tables

# 3. Join the combined identifiers to master in one step
master_itn <- master %>%
  left_join(all_identifiers, by = "ID")

# write_csv(master_itn %>%
#             filter(is.na(Identifier)),
#           file = "~/Documents/ODSi/ODSiData/citation-network/handcheck.csv")

master_auto <- master_itn %>%
  filter(!is.na(Identifier))
#10772

master_handcheck <- read_csv("~/Documents/ODSi/ODSiData/citation-network/handcheck.csv")%>%distinct(ID, .keep_all = TRUE)
#51

master_final <- bind_rows(master_auto, master_handcheck)
#10823
```

# Finalizing Identifying Info

```{r}

bib <- bib2df::bib2df(file = "~/Documents/ODSi/ODSiData/citation-network/Full_Network/full_bib.bib")
```

# Foreign Key

```{r}
citing_dict <- master_final %>%
  mutate(call_id = paste0(citing_paper, "_", local_number))%>%
  left_join(refs_fixed %>%
              select(citing_paper, citing_doi)%>%
              distinct(),
            by = "citing_paper")

set.seed(9134)
citing_dict$id = ids::random_id(n = nrow(citing_dict), bytes = 4)

citing_dict <- citing_dict %>%
  select(id, call_id, citing_paper, citing_doi, local_number, full_citation, Type, TypeOfIdentifier, Identifier)
```

```{r}
linking_dict <- bib %>%
  select(BIBTEXKEY, DOI, PMID, ISSN, ISBN, TITLE, AUTHOR)

linking_dict$ref_id <- ids::random_id(n = nrow(linking_dict), bytes = 4)
```

```{r}
citing_dict2 <- citing_dict %>%
  left_join(linking_dict %>%
              select(citing_bibtex = BIBTEXKEY, DOI, citing_id = ref_id),
            by = c("citing_doi" = "DOI" ))%>%
  select(id, call_id, citing_bibtex, citing_id, local_number, full_citation, Type, TypeOfIdentifier, Identifier)
```

```{r}
linking_long <- linking_dict %>%
  # Select the ID column and the target identifier columns
  select(ref_id, ref_bibtex = BIBTEXKEY, DOI, PMID, ISBN) %>%
  # Pivot the data so column names become a variable and values become a variable
  pivot_longer(
    cols = c(DOI, PMID, ISBN),
    names_to = "TypeOfIdentifier",
    values_to = "Identifier",
    # Ensure all identifiers are treated as characters to avoid type-mismatch errors
    values_transform = list(Identifier = as.character)
  ) %>%
  # Drop empty rows where no identifier existed
  filter(!is.na(Identifier) & Identifier != "") %>%
  # Standardize case: TypeOfIdentifier to uppercase, Identifier to lowercase
  mutate(
    TypeOfIdentifier = toupper(TypeOfIdentifier),
    Identifier = tolower(Identifier)  # <--- FIX ADDED HERE
  )

# 2. Join the reshaped dictionary to your citing dictionary
citing_dict3_linked <- citing_dict2 %>%
  # Standardize the citing_dict2 Identifier to lowercase as well
  mutate(
    Identifier = tolower(as.character(Identifier)) # <--- FIX ADDED HERE
  ) %>%
  # Perform a left join to bring in ref_id where there is a match
  left_join(linking_long, by = c("TypeOfIdentifier", "Identifier"))

unlinked <- citing_dict3_linked %>%
  filter(is.na(ref_id))
```

# Export

-   Zotero library (.bib)

-   citing_dictionary

-   ref_dictionary

```{r}
citing_dictionary <- citing_dict3_linked
# save(citing_dictionary, file = "~/Documents/ODSi/ODSiData/citation-network/Full_Network/citing_dictionary.Rdata")
# write_excel_csv(citing_dictionary, file = "~/Documents/ODSi/ODSiData/citation-network/Full_Network/citing_dictionary.csv" )
```

```{r}
# ref_dictionary <- linking_dict %>%
#   select(ref_id, names(linking_dict)[1:7])
#  save(ref_dictionary, file = "~/Documents/ODSi/ODSiData/citation-network/Full_Network/ref_dictionary.Rdata")
#  write_excel_csv(ref_dictionary, file = "~/Documents/ODSi/ODSiData/citation-network/Full_Network/ref_dictionary.csv" )
```
