---
title: "Batch Correction Attempts"
author: "Sophie Huebler"
format: 
  html:
    code-fold: true
    code-tools: true
    toc: true
    embed-resources: true
editor: visual
execute:
  warning: false
  cache: false
engine: knitr
---

```{r}
#| label: libraries
#| output: false

library(tidyverse)
#BiocManager::install("sva")
library(sva)
library(biomformat)
library(vegan)
library(phyloseq)
library(ggExtra)
library(ComplexHeatmap)
library(ggridges)
```

```{r}
#| label: sourcing
#| output: false


load("~/Documents/ODSi/ODSiData/AnalyticCohort/uncorrected_ps.Rdata")

source("/Users/sophiehuebler/Documents/ODSi/ODSiData/Functions/tree_coloring.R")


```

```{r}
#| label: diversity_functions

permanova_label_generate <- function(perm_res,
                                     ord_res,
                                     x_position = "l",
                                     y_position = "l",
                                     x_perc = .95,
                                     y_perc = .95){
  
# 1. Extract the rounded F and p-value from the adonis2 results
permanova_f <- perm_res$F[1]
permanova_p <- perm_res$`Pr(>F)`[1]


rounded_f <- round(permanova_f, 2)
rounded_p <- round(permanova_p, 3)

# Text
if (permanova_p < 0.001) {
  p_text <- "< 0.001"
} else {
  p_text <- paste0("=", rounded_p)
}

box_label <- paste0("PERMANOVA:\nF=", rounded_f, ", p", p_text)

# 2. Placement
ord_df <- as.data.frame(ord_res$vectors[, 1:2])

if(x_position == "l"){
  coord_x <- min(ord_df$Axis.1)
}else if(x_position == "u"){
    coord_x <- max(ord_df$Axis.1)
  }

if(y_position == "l"){
    coord_y <- min(ord_df$Axis.2)
  }else if(y_position == "u"){
    coord_y <- max(ord_df$Axis.2)
  }


box_x <- coord_x * as.numeric(x_perc)
box_y <- coord_y * as.numeric(y_perc)


return(list("box_label" = box_label,
            "box_x" = box_x,
            "box_y" = box_y))
  
}

beta_pipe <- function(phylo_obj,
                      rare_depth = NULL,
                      ord_method,
                      title_str,
                      norm_method = "rarefy",
                      box_label_positions = c("l", "l"),
                      box_label_percs = c(.95,.95),
                      overide_positions = NULL){
  
  # 1) Normalize
  
  if (norm_method == "rarefy") {
    if (is.null(rare_depth)) stop("Must provide rare_depth when norm_method is 'rarefy'")
    
    norm_phylo <- phyloseq::rarefy_even_depth(phylo_obj,
                                              sample.size = rare_depth,
                                              replace = FALSE)
  } else if (norm_method == "logCPM") {
    # Calculate log2 Counts Per Million with a pseudocount of 1 to handle zeros
    norm_phylo <- phyloseq::transform_sample_counts(phylo_obj, function(x) {
      log2(((x / sum(x)) * 1e6) + 1)
    })
  } else {
    stop("norm_method must be either 'rarefy' or 'logCPM'")
  }
  

  norm_phylo <- prune_taxa(taxa_sums(norm_phylo)>0,
                           norm_phylo)
  norm_phylo <- prune_samples(sample_sums(norm_phylo)>0, 
                              norm_phylo)
  
  # 2) Ordination
  

  if(ord_method %in% c("bray")){
    distance_obj <- phyloseq::distance(norm_phylo, method = ord_method)
  }
  if(ord_method %in% c("unifrac")){
   distance_obj <-  phyloseq::distance(norm_phylo, method = ord_method,
                       weighted = FALSE)
  }
    
    ord_obj <- ordinate(norm_phylo,
                        method = "PCoA", 
                        distance = distance_obj)
    
    perm_res <- adonis2(
      distance_obj ~ study,
      data = data.frame(sample_data(norm_phylo)),
      permutations = 999)
    
    box <- permanova_label_generate(perm_res, ord_obj, 
                                    box_label_positions[1], box_label_positions[2],
                                    as.numeric(box_label_percs[1]), as.numeric(box_label_percs[2]))
    
     if(!is.null(overide_positions)){
   box$box_x <- overide_positions[1]
   box$box_y <- overide_positions[2]
 }
    
    #print(box)
    
  # 3) Plot
    
    base_plot <- plot_ordination(norm_phylo,
                             ord_obj,
                             color = "study") +
      geom_point(size = 2.5, alpha = 0.8) +          
      stat_ellipse(type = "t", level = 0.95, linetype = 2) +
      theme_bw() +                                   
      theme(legend.position = "bottom")+              
      scale_color_manual(name = "Cohort",
                     labels = c("Allo" = "Allozithro",
                                "Fuji" = "Fujimoto",
                                "Liu" = "Liu"),
                     values = c("gold", "#56B4E9", "#76EE00"))+
      geom_label(x = box$box_x, y = box$box_y, label = box$box_label,
             fill = "white", color = "black", fontface = "bold", size = 4) +
      ggtitle(title_str)+
      xlim(-.6, .6)+
      ylim(-.6, .6)

  # Add the marginal distributions using ggExtra
final_plot <- ggMarginal(base_plot, 
                         type = "density",       
                         groupColour = TRUE, 
                         groupFill = TRUE, 
                         alpha = 0.4)            


final_plot

  
}
```

```{r}
#| label: heatmap_functions

plot_heat <- function(phylo_obj, heat_type){
  
  if (taxa_are_rows(phylo_obj)) {
    otu_mat <- as(otu_table(phylo_obj), "matrix")
  } else {
    otu_mat <- t(as(otu_table(phylo_obj), "matrix"))
  }

if(heat_type == "sparsity"){
  binary_mat <- ifelse(otu_mat > 0, 1, 0)
  
  plot <- Heatmap(binary_mat, 
        name = "Presence",
        col = c("white", "steelblue"), 
        show_row_names = FALSE, 
        cluster_columns = TRUE,
        column_title = "Samples",
        row_title = "Taxa (Sparsity)")
  
}

if(heat_type == "size"){

log_mat <- log10(otu_mat + 1)


lib_size <- sample_sums(phylo_obj)


ha = HeatmapAnnotation(LibSize = anno_barplot(lib_size))

plot <- Heatmap(log_mat, 
        name = "Log10(Counts)",
        top_annotation = ha,
        col = viridis::viridis(100),
        show_row_names = FALSE,
        cluster_columns = FALSE, 
        column_title = "Samples")
}
  
  plot

}
```

# Original object

Starting with our analytic_samples_ps object (called here uncorrected_ps):

```{r}
#| label: print_uncorrected_ps

print(uncorrected_ps)
```

## Sparsity patterns

### Sparsity

```{r}
#| label: uncorrected_sparsity_heatmap

plot_heat(uncorrected_ps, "sparsity")
```

### Library size variation

```{r}
#| label: uncorrected_size_heatmap

plot_heat(uncorrected_ps, "size")
```

Subset for only the taxa which appear in all studies:

```{r}
#| label: uncorrected_size_heatmap_shared_taxa

## Showing only ~450 taxa shared by all studies
shared_count_matrix <- as(otu_table(uncorrected_ps), "matrix")
if (!taxa_are_rows(uncorrected_ps)) {
  shared_count_matrix <- t(shared_count_matrix)
}

## Extract batch (study) 
shared_count_meta <- data.frame(sample_data(uncorrected_ps))
shared_count_meta <- shared_count_meta[colnames(shared_count_matrix), , drop = FALSE]

shared_batch_var <- droplevels(factor(shared_count_meta$study))


# Filter for samples that have taxa which only appears in a single batch

## 1) Identify taxa that are non-zero in at least one sample of at least two batches
shared_shared_taxa <-  rowSums(sapply(levels(shared_batch_var), function(b) {
  rowSums(shared_count_matrix[, shared_batch_var == b, drop = FALSE]) > 0
  })) >= 3

shared_ps <- prune_taxa(shared_shared_taxa, uncorrected_ps)
shared_ps <- prune_samples(!(grepl("_", sample_data(shared_ps)$sample.id)),
                         shared_ps)

shared_otu_mat <- t(as(otu_table(shared_ps), "matrix"))

shared_log_mat <- log10(shared_otu_mat + 1)


shared_lib_size <- sample_sums(shared_ps)


shared_ha = HeatmapAnnotation(LibSize = anno_barplot(shared_lib_size))

Heatmap(shared_log_mat, 
        name = "Log10(Counts)",
        top_annotation = shared_ha,
        col = viridis::viridis(100),
        show_row_names = FALSE,
        cluster_columns = FALSE, 
        column_title = "Samples")



```

# ComBat-seq correction (applied to raw counts)

-   Combat seq correction on original counts

-   Not preserving any biological signal

-   Keeping only taxa that are non-zero in at least 1 sample of at least 2 batches

-   Otherwise standard implementation

```{r}
#| label: comBat-seq

## ComBat_seq requires taxa are rows, samples are cols
count_matrix <- as(otu_table(uncorrected_ps), "matrix")
if (!taxa_are_rows(uncorrected_ps)) {
  count_matrix <- t(count_matrix)
}

## Extract batch (study) 
count_meta <- data.frame(sample_data(uncorrected_ps))
count_meta <- count_meta[colnames(count_matrix), , drop = FALSE]

batch_var <- droplevels(factor(count_meta$study))


# Filter for samples that have taxa which only appears in a single batch

## 1) Identify taxa that are non-zero in at least one sample of at least two batches
shared_taxa <-  rowSums(sapply(levels(batch_var), function(b) {
  rowSums(count_matrix[, batch_var == b, drop = FALSE]) > 0
  })) >= 2

## 2) Filter count matrix for only those taxa shared between at least 2 studies
shared_count_matrix <- count_matrix[shared_taxa, , drop = FALSE]

## 3) Identify samples that do not have any any shared taxa in library
bad_samples <- colnames(shared_count_matrix)[colSums(shared_count_matrix) == 0]
#print(bad_samples)

## 4) Identify samples that should be kept
sample_keep <- colSums(shared_count_matrix) > 0

## 5) Filter count matrix, batch var, and bio var
count_matrix_keep <- count_matrix[, sample_keep]
count_meta_keep <- count_meta[sample_keep,]
batch_var_keep <- batch_var[sample_keep]

adjusted_counts2 <- sva::ComBat_seq(
  counts = count_matrix_keep,
  batch  = batch_var_keep,
  group  = NULL
)


combatseq_ps <- prune_samples(!(sample_data(uncorrected_ps)$sample.id %in% bad_samples), uncorrected_ps)
otu_table(combatseq_ps) <- otu_table(adjusted_counts2, taxa_are_rows = TRUE)
```

```{r}
#| label: print_corrected_ps

print(combatseq_ps)

```

## Sparsity patterns

### Sparsity

```{r}
#| label: corrected_sparsity_heatmap

plot_heat(combatseq_ps, "sparsity")
```

### Library size variation

```{r}
#| label: corrected_size_heatmap

plot_heat(combatseq_ps, "size")
```

# Ridge Plots

```{r}
get_melted <- function(ps, label) {
  psmelt(ps) %>%
    mutate(source = label)
}

orig <- get_melted(uncorrected_ps, "Before Correction")
corr <- get_melted(combatseq_ps, "After Correction")
combined <- bind_rows(orig, corr) %>%
  mutate(source = factor(source, levels = c("Before Correction", "After Correction")))
```

## Per study total read count before/after

```{r}
#| label: ridge plots 1
#| output: false

fig1_data <- combined %>%
  group_by(Sample, study, source) %>%
  summarise(total_reads = sum(Abundance), .groups = "drop") %>%
  mutate(log_total = log10(total_reads + 1))

fig1 <- ggplot(fig1_data, aes(x = log_total, y = factor(study), fill = study)) +
  geom_density_ridges(alpha = 0.65, scale = 1.2, rel_min_height = 0.01) +
  facet_wrap(~source) +
  theme_ridges(grid = FALSE) +
  theme(#legend.position = "none",
        strip.text = element_text(face = "bold")) +
  labs(
    title    = "Per-Sample Total Read Count Distribution by Study",
    x        = "log10(Total Read Count + 1)",
    y        = "Study"
  )+              
      scale_fill_manual(name = "Cohort",
                     labels = c("Allo" = "Allozithro",
                                "Fuji" = "Fujimoto",
                                "Liu" = "Liu"),
                     values = c("gold", "#56B4E9", "#76EE00"))



```

```{r}
print(fig1)
```

## Top 25 taxa abundance distribution across samples

```{r}
#| label: ridge plot 2
#| output: false

N_TAXA <- 25  # adjust as needed
COLLAPSE_RANK <- "Genus"  # change to "Family", "Phylum", etc. as appropriate

# Collapse to desired rank
ps_orig_genus <- tax_glom(uncorrected_ps, taxrank = COLLAPSE_RANK, NArm = FALSE)
ps_corr_genus <- tax_glom(combatseq_ps, taxrank = COLLAPSE_RANK, NArm = FALSE)

# Get top N taxa by mean abundance from original
top_taxa_names <- names(sort(
  rowMeans(t(otu_table(ps_orig_genus))),
  decreasing = TRUE
))[1:N_TAXA]

# Subset both objects to those taxa
ps_orig_top <- prune_taxa(top_taxa_names, ps_orig_genus)
ps_corr_top <- prune_taxa(top_taxa_names, ps_corr_genus)

# Melt and combine
fig2_orig <- psmelt(ps_orig_top) %>% mutate(source = "Before Correction")
fig2_corr <- psmelt(ps_corr_top) %>% mutate(source = "After Correction")
fig2_data <- bind_rows(fig2_orig, fig2_corr) %>%
  mutate(
    source     = factor(source, levels = c("Before Correction", "After Correction")),
    log_abund  = log10(Abundance + 1),
    taxon_label = coalesce(get(COLLAPSE_RANK), OTU)  # pulls the genus/family/etc. column
  ) %>%
  # Order taxa by median abundance for readability
  group_by(taxon_label) %>%
  mutate(median_abund = median(log_abund)) %>%
  ungroup() %>%
  mutate(taxon_label = fct_reorder(taxon_label, median_abund))%>%
  select(OTU, taxon_label, study, Sample, source, log_abund)

fig2 <- ggplot(fig2_data, aes(x = log_abund, y = taxon_label, color = study)) +
  geom_density_ridges(alpha = 0.05, scale = 1.5, rel_min_height = 0.01) +
  facet_wrap(~source) +
  # scale_fill_manual(values = c("Before Correction" = "#E07B54",
  #                              "After Correction"  = "#5B8DB8")) +
  theme_ridges(grid = FALSE) +
  theme(#legend.position = "none",
        axis.text.y     = element_text(size = 8),
        strip.text      = element_text(face = "bold")) +
  labs(
    title    = paste0("Figure 2: Top ", N_TAXA, " Taxa Abundance Distributions"),
    x        = "log10(Abundance + 1)",
    y        = COLLAPSE_RANK
  )+              
      scale_color_manual(name = "Cohort",
                     labels = c("Allo" = "Allozithro",
                                "Fuji" = "Fujimoto",
                                "Liu" = "Liu"),
                     values = c("gold", "#56B4E9", "#76EE00"))




```

```{r}
print(fig2)
```

## Per study read count before/after

```{r}
#| label: ridge plot 3
#| output: false


get_sample_totals <- function(ps, label) {
  otu <- as.data.frame(otu_table(ps))
  if (!taxa_are_rows(ps)) otu <- t(otu)
  
  data.frame(
    SampleID   = colnames(otu),
    total_reads = colSums(otu),
    source      = label
  ) %>%
    left_join(
      data.frame(sample_data(ps)) %>% rownames_to_column("SampleID"),
      by = "SampleID"
    )
}

orig_totals <- get_sample_totals(uncorrected_ps,  "Before")
corr_totals <- get_sample_totals(combatseq_ps, "After")

# Join and compute difference and fold change
delta_df <- orig_totals %>%
  select(SampleID, study, before = total_reads) %>%   # replace Study with your variable name
  left_join(
    corr_totals %>% select(SampleID, after = total_reads),
    by = "SampleID"
  ) %>%
  mutate(
    delta      = after - before,                       
    log2fc     = log2((after + 1) / (before + 1)),     
    pct_change = 100 * (after - before) / (before + 1)  
  ) %>%
  group_by(study) %>%
  mutate(median_abs_delta = median(abs(delta))) %>%
  ungroup() %>%
  mutate(study = fct_reorder(study, median_abs_delta))

plot_a <- ggplot(delta_df, aes(x = delta, y = study, fill = study)) +
  geom_density_ridges(
    alpha          = 0.75,
    scale          = 1.3,
    rel_min_height = 0.01,
    quantile_lines = TRUE,   # adds median line
    quantiles      = 2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray30", linewidth = 0.7) +
  scale_fill_viridis_d(option = "turbo", guide = "none") +
  theme_ridges(grid = FALSE) +
  theme(
    axis.text.y = element_text(size = 10),
    plot.subtitle = element_text(size = 9, color = "gray40")
  ) +
  labs(
    title    = "Read Count Change per Sample by Study",
    subtitle = "After − Before correction | Dashed line = no change | Tick = median",
    x        = "Read Count Difference (After − Before)",
    y        = "Study"
  )+              
      scale_fill_manual(name = "Cohort",
                     labels = c("Allo" = "Allozithro",
                                "Fuji" = "Fujimoto",
                                "Liu" = "Liu"),
                     values = c("gold", "#56B4E9", "#76EE00"))

```

```{r}
#| label: print ridge plot 3a 




plot_a
```

```{r}
#| label: ridge plot 3b
#| output: false


plot_b <- ggplot(delta_df, aes(x = log2fc, y = study, fill = study)) +
  geom_density_ridges(
    alpha          = 0.75,
    scale          = 1.3,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles      = 2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray30", linewidth = 0.7) +
  scale_fill_viridis_d(option = "turbo", guide = "none") +
  theme_ridges(grid = FALSE) +
  theme(
    axis.text.y   = element_text(size = 10),
    plot.subtitle = element_text(size = 9, color = "gray40")
  ) +
  labs(
    title    = "Log2 Fold Change per Sample by Study",
    subtitle = "log2(After + 1 / Before + 1) | Dashed line = no change | Tick = median",
    x        = "Log2 Fold Change",
    y        = "Study"
  )+              
      scale_fill_manual(name = "Cohort",
                     labels = c("Allo" = "Allozithro",
                                "Fuji" = "Fujimoto",
                                "Liu" = "Liu"),
                     values = c("gold", "#56B4E9", "#76EE00"))


```

```{r}
#| label: print ridge plot 3b


plot_b
```

```{r}
#| label: ridge plot 3c
#| output: false
#| 
plot_c <- ggplot(delta_df, aes(x = pct_change, y = study, fill = study)) +
  geom_density_ridges(
    alpha          = 0.75,
    scale          = 1.3,
    rel_min_height = 0.01,
    quantile_lines = TRUE,
    quantiles      = 2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray30", linewidth = 0.7) +
  scale_fill_viridis_d(option = "turbo", guide = "none") +
  theme_ridges(grid = FALSE) +
  theme(
    axis.text.y   = element_text(size = 10),
    plot.subtitle = element_text(size = 9, color = "gray40")
  ) +
  labs(
    title    = "Percent Read Count Change per Sample by Study",
    subtitle = "100 × (After − Before) / Before | Dashed line = no change | Tick = median",
    x        = "Percent Change (%)",
    y        = "Study"
  )+              
      scale_fill_manual(name = "Cohort",
                     labels = c("Allo" = "Allozithro",
                                "Fuji" = "Fujimoto",
                                "Liu" = "Liu"),
                     values = c("gold", "#56B4E9", "#76EE00"))

```

```{r}
#| label: print ridge plot 3c

plot_c
```

# Comparing beta diversity before/after

## Bray-Curtis

### Rarefied

-   Rarefaction applied after correction

```{r}
#| label: bray_rare
#| output: false

uncorrected_rare900_beta_bray <- beta_pipe(uncorrected_ps,
          900,
          "bray",
"BEFORE:
Beta Diversity (Bray-Curtis)
Uncorrected
Rarefied to 900",
"rarefy",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

combatseq_rare900_beta_bray <- beta_pipe(combatseq_ps,
          900,
          "bray",
"AFTER:
Beta Diversity (Bray-Curtis)
ComBat-seq corrected
Rarefied to 900",
"rarefy",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))


```

```{r}
#| label: print_bray_rare_before

uncorrected_rare900_beta_bray

```

```{r}
#| label: print_bray_rare_after

combatseq_rare900_beta_bray

```

### Log counts per million

```{r}
#| label: bray_logCPM
#| output: false

uncorrected_logCPM_beta_bray <- beta_pipe(uncorrected_ps,
          NULL,
          "bray",
"BEFORE:
Beta Diversity (Bray-Curtis)
Uncorrected
Log CPM",
"logCPM",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

combatseq_logCPM_beta_bray <- beta_pipe(combatseq_ps,
          900,
          "bray",
"AFTER:
Beta Diversity (Bray-Curtis)
ComBat-seq corrected
Log CPM",
"logCPM",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))


```

```{r}
#| label: print_bray_logCPM_before

uncorrected_logCPM_beta_bray

```

```{r}
#| label: print_bray_logCPM_after

combatseq_logCPM_beta_bray

```

## Unweighted Unifrac

### Rarefied

-   Rarefaction applied after correction

```{r}
#| label: unifrac_rare
#| output: false

uncorrected_rare900_beta_unifrac <- beta_pipe(uncorrected_ps,
          900,
          "unifrac",
"BEFORE:
Beta Diversity (Unweighted Unifrac)
Uncorrected
Rarefied to 900",
"rarefy",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

combatseq_rare900_beta_unifrac <- beta_pipe(combatseq_ps,
          900,
          "unifrac",
"AFTER:
Beta Diversity (Unweighted Unifrac)
ComBat-seq corrected
Rarefied to 900",
"rarefy",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))


```

```{r}
#| label: print_unifrac_rare_before

uncorrected_rare900_beta_unifrac

```

```{r}
#| label: print_unifrac_rare_after

combatseq_rare900_beta_unifrac

```

### Log counts per million

```{r}
#| label: unifrac_logCPM
#| output: false

uncorrected_logCPM_beta_unifrac <- beta_pipe(uncorrected_ps,
          NULL,
          "unifrac",
"BEFORE:
Beta Diversity (Unweighted Unifrac)
Uncorrected
Log CPM",
"logCPM",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

combatseq_logCPM_beta_unifrac <- beta_pipe(combatseq_ps,
          900,
          "unifrac",
"AFTER:
Beta Diversity (Unweighted Unifrac)
ComBat-seq corrected
Log CPM",
"logCPM",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))


```

```{r}
#| label: print_unifrac_logCPM_before

uncorrected_logCPM_beta_unifrac

```

```{r}
#| label: print_unifrac_logCPM_after

combatseq_logCPM_beta_unifrac

```
