---
title: "Merging_Metadata"
format: html
editor: visual
---

# Set up

```{r}
library(tidyverse)
```

# Target schema

-   Identifying information: sample-id, study, person, and timepoint of the sample.

-   Demographics: age, sex, disease

-   Outcome: gut-gvhd, death, relpse

# Reading in data

```{r}
source("~/Documents/ODSi/ODSiData/Merging/Harmonization/harmonize_gvhd_metadata.R")
```

Each block reads a study's parsed metadata (produced by that study's `Parsing_metadata.qmd`) and maps its columns into the target schema. Comments note what the study does and does NOT provide.

```{r}
artacho_raw <- ensure_sample_id(read_tsv("~/Documents/ODSi/ODSiData/Artacho2024/Metadata/artacho_meta_qiime.tsv"))
damico_raw <- ensure_sample_id(read_tsv("~/Documents/ODSi/ODSiData/DAmico2019/Metadata/damico_meta_qiime.tsv"))
fujimoto_raw <- ensure_sample_id(read_tsv("~/Documents/ODSi/ODSiData/Fujimoto2024/Metadata/fuji_meta_qiime.tsv"))
ingham_raw <- ensure_sample_id(read_tsv("~/Documents/ODSi/ODSiData/Ingham2019/Metadata/ingham_meta_qiime.tsv"))
liu_raw  <- ensure_sample_id(read_tsv("~/Documents/ODSi/ODSiData/Liu2017/Metadata/liu_meta_qiime.tsv")) 
vallet_raw <- ensure_sample_id(read_tsv("~/Documents/ODSi/ODSiData/Vallet2023/Metadata/vallet_meta_qiime.tsv"))

jarosch_raw <- read_csv("~/Documents/ODSi/ODSiData/Jarosch2023/Metadata/Raw_metadata/meta_samples.csv")%>%
  dplyr::rename(
    `sample-id` = Run
  )
```

## Person Crosswalk

```{r}
person_crosswalk <- bind_rows(
    make_person_crosswalk(
      artacho_raw[["Patient"]], study_labels[["artacho"]],
      person_prefixes[["artacho"]], "Patient"
    ),
    make_person_crosswalk(
      damico_raw[["Patient"]], study_labels[["damico"]],
      person_prefixes[["damico"]], "Patient"
    ),
    make_person_crosswalk(
      fujimoto_raw[["Patient"]], study_labels[["fujimoto"]],
      person_prefixes[["fujimoto"]], "Patient"
    ),
    make_person_crosswalk(
      ingham_raw[["patient"]], study_labels[["ingham"]],
      person_prefixes[["ingham"]], "patient"
    ),
    make_person_crosswalk(
      liu_raw[["host_subject_id"]], study_labels[["liu"]],
      person_prefixes[["liu"]], "host_subject_id"
    ),
    make_person_crosswalk(
      vallet_raw[["allozithro_id"]], study_labels[["vallet"]],
      person_prefixes[["vallet"]], "allozithro_id"
    )
  )
```

# Harmonizing

```{r}
source("~/Documents/ODSi/ODSiData/Merging/Harmonization/harmonize_gvhd_metadata.R")
```

```{r}
artacho <- harmonize_artacho(artacho_raw, person_crosswalk)
damico <- harmonize_damico(damico_raw, person_crosswalk)
fujimoto <- harmonize_fujimoto(fujimoto_raw, person_crosswalk)
ingham <- harmonize_ingham(ingham_raw, person_crosswalk)
liu <- harmonize_liu(liu_raw, person_crosswalk)
vallet <- harmonize_vallet(vallet_raw, person_crosswalk)
```

```{r}
jarosch <- jarosch_raw %>%
  select(`sample-id`)%>%
  mutate(study = "Jarosch")%>%
  to_schema()
```

```{r}
harmonized_meta <- bind_rows(artacho,
                             damico,
                             fujimoto,
                             ingham,
                             jarosch,
                             liu,
                             vallet)
```

## Setting up final data frame

```{r}
write_tsv(harmonized_meta,
          file = "~/Documents/ODSi/ODSiData/Merging/merged_metadata.tsv")
```
