---
title: "Merging_Metadata"
format: html
editor: visual
---

# Set up

```{r}
library(tidyverse)
```

# Target schema

The merged frame has `sample-id` as the first column, then one row per sample across all 7 studies. Not every study populates every column; anything a study does not measure is left `NA`. Values are kept as-is (raw) here — harmonization happens in the section below.

Everything is coerced to character during assembly so that raw codings from the different studies (e.g. Liu's "yes"/"no" relapse vs. Ingham's 0/1) can coexist in one column without type clashes. The harmonization step will re-type each column.

```{r}
target_cols <- c(
  "sample-id",
  "study",
  "person",
  "sex",
  "age",
  "disease",
  "gut_gvhd",
  "gut_gvhd_stage",
  "transplant_day",
  "sample_day",
  "sample_time_rel_transplant",
  "gut_gvhd_time_rel_transplant",
  "death",
  "cause_of_death",
  "death_time_rel_transplant",
  "relapse",
  "relapse_time_rel_transplant"
)

# Coerce to character, add any missing target columns as NA, order columns.
to_schema <- function(df) {
  df <- df %>% mutate(across(everything(), as.character))
  for (col in target_cols) {
    if (!col %in% names(df)) df[[col]] <- NA_character_
  }
  df[target_cols]
}
```

# Reading in data

Each block reads a study's parsed metadata (produced by that study's `Parsing_metadata.qmd`) and maps its columns into the target schema. Comments note what the study does and does NOT provide.

## Liu2017

Overall aGVHD only (`agvhd` / `agvhd_organ`), no gut-specific staging, so `gut_gvhd*` are left blank. Donors have no disease -\> set to "none".

```{r}
liu <- read_tsv("~/Documents/ODSi/ODSiData/Liu2017/Metadata/liu_meta_qiime.tsv") 
  # transmute(
  #   `sample-id` = ID,
  #   study       = "Liu2017",
  #   person      = host_subject_id,
  #   sex         = sex,
  #   age         = age,
  #   disease     = if_else(donor_or_patient == "Donor", "none", disease),
  #   death       = deceased,
  #   relapse     = relapsed
  # ) 
```

## Fujimoto2024

Overall aGVHD only (`AGVHD`), no gut-specific staging. `Timepoint` is a sampling label ("pre", "day0", "day14", ...); parse the numeric day relative to transplant ("pre" -\> NA).

```{r}
fuji <- read_tsv("~/Documents/ODSi/ODSiData/Fujimoto2024/Metadata/fuji_meta_qiime.tsv") #%>%
  # transmute(
  #   `sample-id`                = `sample-id`,
  #   study                      = "Fujimoto2024",
  #   person                     = Patient,
  #   sex                        = Sex,
  #   age                        = Age,
  #   disease                    = Disease,
  #   sample_time_rel_transplant = suppressWarnings(as.numeric(str_remove(Timepoint, "day")))
  # ) %>%
  # to_schema()
```

## Artacho2024

Has gut-specific GVHD grade (`gut_gvhd_grade`). `Timepoint` is days relative to transplant.

```{r}
artacho <- read_tsv("~/Documents/ODSi/ODSiData/Artacho2024/Metadata/artacho_meta_qiime.tsv") #%>%
  # transmute(
  #   `sample-id`                = `sample-id`,
  #   study                      = "Artacho2024",
  #   person                     = Patient,
  #   sex                        = Gender,
  #   age                        = Age,
  #   disease                    = Disease,
  #   gut_gvhd                   = case_when(is.na(gut_gvhd_grade) ~ NA_real_,
  #                                          gut_gvhd_grade > 0     ~ 1,
  #                                          TRUE                   ~ 0),
  #   gut_gvhd_stage             = gut_gvhd_grade,
  #   sample_time_rel_transplant = Timepoint
  # ) %>%
  # to_schema()
```

## Ingham2019

No gut-specific GVHD (overall `agvhd`/`agvhd_grade` only), so `gut_gvhd*` blank. Rich outcome data: death, cause of death, time-to-event, relapse. Dates are kept as raw strings; `timepoint` is days post-transplant. Note `sex` is coded 1/2 and `age` carries decimals — harmonize later.

```{r}
ingham <- read_tsv("~/Documents/ODSi/ODSiData/Ingham2019/Metadata/ingham_meta_qiime.tsv") #%>%
  # transmute(
  #   `sample-id`                 = `sample-id`,
  #   study                       = "Ingham2019",
  #   person                      = patient,
  #   sex                         = sex,
  #   age                         = age,
  #   disease                     = disease,
  #   transplant_day              = transplant_date,
  #   sample_day                  = sampling_date,
  #   sample_time_rel_transplant  = timepoint,
  #   death                       = death,
  #   cause_of_death              = cause_of_death,
  #   death_time_rel_transplant   = tte_death,
  #   relapse                     = relapse,
  #   relapse_time_rel_transplant = tte_relapse
  # ) %>%
  # to_schema()
```

## Vallet2023

Overall aGVHD only, so `gut_gvhd*` blank. Has transplant date (`dat.hsct`) and sample date (`sample_date`); compute sample time relative to transplant in days.

```{r}
vallet <- read_tsv("~/Documents/ODSi/ODSiData/Vallet2023/Metadata/vallet_meta_qiime.tsv") #%>%
  # transmute(
  #   `sample-id`                = `sample-id`,
  #   study                      = "Vallet2023",
  #   person                     = individual,
  #   sex                        = sex,
  #   age                        = age,
  #   disease                    = diagnosis2,
  #   transplant_day             = dat.hsct,
  #   sample_day                 = sample_date,
  #   sample_time_rel_transplant = as.numeric(as.Date(sample_date) - as.Date(dat.hsct)),
  #   death                      = death,
  #   relapse                    = relapse
  # ) %>%
  #  to_schema()
```

## DAmico2019

Has gut-specific GVHD grade (`gut_gvhd_grade`) and GVHD onset day (`gvhd_day`). `Timepoint` is days relative to transplant. No death/relapse fields captured.

```{r}
damico <- read_tsv("~/Documents/ODSi/ODSiData/DAmico2019/Metadata/damico_meta_qiime.tsv") #%>%
  # transmute(
  #   `sample-id`                    = `sample-id`,
  #   study                          = "DAmico2019",
  #   person                         = Patient,
  #   sex                            = Sex,
  #   age                            = Age,
  #   disease                        = Diagnosis,
  #   gut_gvhd                       = case_when(is.na(gut_gvhd_grade) ~ NA_real_,
  #                                              gut_gvhd_grade > 0     ~ 1,
  #                                              TRUE                   ~ 0),
  #   gut_gvhd_stage                 = gut_gvhd_grade,
  #   sample_time_rel_transplant     = Timepoint,
  #   gut_gvhd_time_rel_transplant   = gvhd_day
  # ) %>%
  # to_schema()
```

## Jarosch2023

Clinical data is not yet matched to the sequenced samples, so only `sample-id` and `study` are populated; everything else stays blank. Sample IDs come from the raw sample sheet (the ENA run accessions).

```{r}
jarosch <- read_csv("~/Documents/ODSi/ODSiData/Jarosch2023/Metadata/Raw_metadata/meta_samples.csv") %>%
  dplyr::rename(
    `sample-id` = Run
  )
```

## Setting up final data frame

```{r}
merged_metadata <- bind_rows(
  liu, fuji, artacho, ingham, vallet, damico, jarosch
)

# sanity checks
merged_metadata %>% count(study)
nrow(merged_metadata)
```

```{r}
write_tsv(merged_metadata,
          file = "~/Documents/ODSi/ODSiData/Merging/merged_metadata.tsv")
```

# Harmonizing
