---
title: "Untitled"
format: html
editor: visual
---

# Libraries

```{r}
library(phyloseq)
library(qiime2R)
library(biomformat)
library(readr)
library(tidyverse)
library(ape)
library(Biostrings)
library(ggtree)
```

# Files

```{r}
table_qza <- "~/Documents/ODSi/ODSiData/Merging/full_cohort/merged-table.qza"
taxonomy_qza  <- "~/Documents/ODSi/ODSiData/Merging/full_cohort/taxonomy-gg.qza"
repseqs_qza   <- "~/Documents/ODSi/ODSiData/Merging/full_cohort/merged-seqs.qza"      
tree_nwk     <- "~/Documents/ODSi/ODSiData/Merging/full_cohort/merged-tree.nwk"
meta_tsv <- "~/Documents/ODSi/ODSiData/Merging/merged_metadata.tsv"
```

```{r}
qt_orig <- read_qza(table_qza)$data
tax_q  <- read_qza(taxonomy_qza)$data
tree    <- ape::read.tree(tree_nwk)
rep_seqs <- tryCatch(read_qza(repseqs_qza)$data, error = function(e) NULL)
meta <- read_tsv(meta_tsv)
```

# Phyloseq

## OTU

```{r}
otu_mat <- otu_table(qt_orig, taxa_are_rows = TRUE)
```

## Taxonomy

```{r}
parse_taxonomy <- function(tax_df) {
  # tax_df expected to have columns: Feature.ID (or similar) and Taxon (string "k__; p__; ...")
  cnames <- colnames(tax_df)
  idcol <- cnames[1]    # usually "Feature.ID"
  taxcol <- cnames[2]   # usually "Taxon"
  tax_df2 <- as.data.frame(tax_df, stringsAsFactors = FALSE)
  colnames(tax_df2)[1:2] <- c("FeatureID","Taxon")
  tax_df2$Taxon <- as.character(tax_df2$Taxon)
  ranks <- c("Kingdom","Phylum","Class","Order","Family","Genus","Species")
  tax_sep <- tidyr::separate(tax_df2, Taxon, into = ranks, sep = ";\\s*", fill = "right", remove = FALSE)
  clean_rank <- function(x) {
    x <- ifelse(is.na(x) | x == "" , NA, x)
    x <- gsub("^[dkpcofgs]__|^__", "", x)   # remove prefixes like 'k__'
    trimws(x)
  }
  tax_mat <- tax_sep %>%
    dplyr::select(all_of(ranks)) %>%
    mutate_all(clean_rank) %>%
    as.matrix()
  rownames(tax_mat) <- tax_sep$FeatureID
  tax_mat
}

```

```{r}
tax_mat <- parse_taxonomy(tax_q)
tax_mat <- tax_table(tax_mat)
```

## Seqs

```{r}
seqs <- Biostrings::DNAStringSet(rep_seqs)
names(seqs) <- names(rep_seqs)
```

## Meta

```{r}

setdiff(meta$`sample-id`, colnames(otu_mat))->excluded
meta_df <- sample_data(meta %>%
                         filter(!(`sample-id` %in% excluded)))
rownames(meta_df)<- meta_df$`sample-id`
```

## Combining

```{r}
print("Matching otu rownames and tree tip labels:")
all(rownames(otu_mat) %in% tree$tip.label)

print(sprintf("OTU features: %d ; tree tips: %d", ntaxa(otu_mat), length(tree$tip.label)))
print(sprintf("Intersection: %d", sum(rownames(otu_mat) %in% tree$tip.label)))


print(sprintf("OTU sample: %d ; meta samples: %d", ncol(otu_mat), nrow(meta_df)))
print(sprintf("Intersection: %d", sum(colnames(otu_mat) %in% rownames(meta_df))))
```

```{r}
ps <- phyloseq(otu_mat, tax_mat, meta_df, phy_tree(tree), refseq(seqs))
print(ps)
```

# Visualizing

```{r}
source("/Users/sophiehuebler/Documents/ODSi/ODSiData/Functions/color_tips.R")
```

```{r}
circular_tree <- ggtree(ps@phy_tree, layout = "circular", branch.length = "none") +
  geom_tippoint( size = 1.6, alpha = 0.9)

circular_tree
```

```{r}
genus_tree <- vizualize_collapse(ps, "Genus", show_labels = TRUE)
```

# Beta diversity

```{r}
library(sva)
library(biomformat)
library(vegan)
library(phyloseq)
library(ggExtra)
```

```{r}
permanova_label_generate <- function(perm_res,
                                     ord_res,
                                     x_position = "l",
                                     y_position = "l",
                                     x_perc = .95,
                                     y_perc = .95){
  
  # 1. Extract the F and p-value from the adonis2 results
permanova_f <- perm_res$F[1]
permanova_p <- perm_res$`Pr(>F)`[1]

# Create the formatted text label for the box
# and format p-values less than 0.001 as "p < 0.001"
# For this example, let's just make it look like the plot you showed.
# We'll need some logic, as p=0.001 might be rounded from 0.0007,
# while p=0.0001 might be rounded from 0.00006.
# We'll use a rounded number to 3 decimal places as a starting point.
rounded_f <- round(permanova_f, 2)
rounded_p <- round(permanova_p, 3)

# Then we create the text label, adding logic to show < 0.001 if needed.
if (permanova_p < 0.001) {
  p_text <- "< 0.001"
} else {
  p_text <- paste0("=", rounded_p)
}

box_label <- paste0("PERMANOVA:\nF=", rounded_f, ", p", p_text)

# We need to decide where to place the box. We will place it based on the
# coordinates in the plot, which means we have to find out what the maximum
# x and y coordinates are. We'll find them from the ordination object.
# Find the range of PCoA 1 and PCoA 2
ord_df <- as.data.frame(ord_res$vectors[, 1:2])

if(x_position == "l"){
  coord_x <- min(ord_df$Axis.1)
}else if(x_position == "u"){
    coord_x <- max(ord_df$Axis.1)
  }

if(y_position == "l"){
    coord_y <- min(ord_df$Axis.2)
  }else if(y_position == "u"){
    coord_y <- max(ord_df$Axis.2)
  }



# Set the corner position of the label relative to the plot.
# We'll use 90% of the max range for each axis.
# Adjust these values based on your specific plot data.
box_x <- coord_x * as.numeric(x_perc)
box_y <- coord_y * as.numeric(y_perc)


return(list("box_label" = box_label,
            "box_x" = box_x,
            "box_y" = box_y))
  
}
```

```{r}
beta_pipe <- function(phylo_obj,
                      rare_depth = NULL,
                      ord_method,
                      title_str,
                      norm_method = "rarefy",
                      box_label_positions = c("l", "l"),
                      box_label_percs = c(.95,.95),
                      overide_positions = NULL){
  
  # 1) Normalize
  
  if (norm_method == "rarefy") {
    if (is.null(rare_depth)) stop("Must provide rare_depth when norm_method is 'rarefy'")
    
    norm_phylo <- phyloseq::rarefy_even_depth(phylo_obj,
                                              sample.size = rare_depth,
                                              replace = FALSE)
    
     norm_phylo <- prune_taxa(taxa_sums(norm_phylo)>0,
                             norm_phylo)
    norm_phylo <- prune_samples(sample_sums(norm_phylo)>0,
                                norm_phylo)
  } else if (norm_method == "logCPM") {
    # Calculate log2 Counts Per Million with a pseudocount of 1 to handle zeros
    norm_phylo <- phyloseq::transform_sample_counts(phylo_obj, function(x) {
      log2(((x / sum(x)) * 1e6) + 1)
    })
      
    norm_phylo <- prune_taxa(taxa_sums(norm_phylo)>0,
                             norm_phylo)
    norm_phylo <- prune_samples(sample_sums(norm_phylo)>0,
                                norm_phylo)
    
  } else if(norm_method == "identity"){
    norm_phylo <- phylo_obj
    print("No normalization used, make sure that supplied phylo_obj has normalized relative abundance data")
    if(ord_method != "euclidean"){
    norm_phylo <- prune_taxa(taxa_sums(norm_phylo)>0,
                             norm_phylo)
    norm_phylo <- prune_samples(sample_sums(norm_phylo)>0,
                                norm_phylo)
    }
    
  }else if(norm_method == "clr"){
    # Centered log-ratio (CLR) transform with a naive pseudocount of 1.
    # Extract counts as a samples (rows) x taxa (cols) matrix so we can
    # apply the CLR one sample at a time, then rebuild a phyloseq object
    # that keeps the original sample_data() and tax_table() attached.
    otu <- as(otu_table(phylo_obj), "matrix")
    if (taxa_are_rows(phylo_obj)) {
      otu <- t(otu)   # ensure samples are rows, taxa are columns
    }

    imputed <- otu + 1                                   # naive zero imputation
    clr_mat <- t(apply(imputed, 1,
                       function(r) as.numeric(compositions::clr(r))))
    rownames(clr_mat) <- rownames(imputed)
    colnames(clr_mat) <- colnames(imputed)

    clr_otu <- otu_table(clr_mat, taxa_are_rows = FALSE)
    norm_phylo <- phyloseq(clr_otu,
                           sample_data(phylo_obj),
                           tax_table(phylo_obj))
  } else{
    stop("norm_method must be either 'rarefy' or 'logCPM' or 'identity' or 'clr")
  }
  
  # 2) Ordination
  

  if(norm_method == "clr"){
    # CLR data lives in Aitchison (Euclidean) space, so ordination is PCA with
    # Euclidean distance. PCoA on a Euclidean distance matrix is equivalent to
    # PCA and keeps the downstream ord object structure ($vectors / Axis.1 /
    # Axis.2) intact for permanova_label_generate() and plot_ordination().
    if(!missing(ord_method) && !is.null(ord_method) && ord_method != "euclidean"){
      warning("norm_method = 'clr' requires PCA with Euclidean distance; the ",
              "supplied ord_method = '", ord_method, "' is incompatible and is ",
              "being ignored. Using Euclidean distance instead.")
    }
    distance_obj <- phyloseq::distance(norm_phylo, method = "euclidean")
  } else {
    if(ord_method %in% c("bray")){
      distance_obj <- phyloseq::distance(norm_phylo, method = ord_method)
    }
    if(ord_method %in% c("unifrac")){
     distance_obj <-  phyloseq::distance(norm_phylo, method = ord_method,
                         weighted = FALSE)
    }
    if(ord_method %in% c("wunifrac")){
     distance_obj <-  phyloseq::distance(norm_phylo, method = "unifrac",
                         weighted = TRUE)
    }
    if(ord_method %in% c("euclidean")){
     distance_obj <- phyloseq::distance(norm_phylo, method = "euclidean")
    }
  }

    ord_obj <- ordinate(norm_phylo,
                        method = "PCoA",
                        distance = distance_obj)
    
    perm_res <- adonis2(
      distance_obj ~ study,
      data = data.frame(sample_data(norm_phylo)),
      permutations = 999)
    
    box <- permanova_label_generate(perm_res, ord_obj, 
                                    box_label_positions[1], box_label_positions[2],
                                    as.numeric(box_label_percs[1]), as.numeric(box_label_percs[2]))
    
     if(!is.null(overide_positions)){
   box$box_x <- overide_positions[1]
   box$box_y <- overide_positions[2]
 }
    
    #print(box)
    
  # 3) Plot

    # Per-study colors come from the shared palette in
    # Shared_aesthetics/Study_colors.R (`study_colors`, one Okabe-Ito hue per
    # cohort). It is normally in scope already because color_tips.R (sourced in
    # the Visualizing section) sources it; source defensively so beta_pipe still
    # works if this chunk is run on its own.
    if (!exists("study_colors")) {
      source("/Users/sophiehuebler/Documents/ODSi/ODSiData/Shared_aesthetics/Study_colors.R")
    }

    # study_colors is keyed by full "StudyYYYY" names (e.g. "Artacho2024"), but
    # the merged metadata abbreviates most cohorts ("Artacho") while keeping
    # "Liu2017". Match each study value present in this object to the shared name
    # it is a prefix of, so the palette maps onto the actual factor levels rather
    # than silently dropping to NA on a key mismatch.
    present_studies <- sort(unique(as.character(sample_data(norm_phylo)$study)))
    cohort_colors <- vapply(present_studies, function(s) {
      hit <- names(study_colors)[startsWith(names(study_colors), s)]
      if (length(hit) < 1L) {
        stop("No Shared_aesthetics color found for study '", s, "'")
      }
      unname(study_colors[hit[1L]])
    }, character(1))
    names(cohort_colors) <- present_studies

    base_plot <- plot_ordination(norm_phylo,
                             ord_obj,
                             color = "study") +
      geom_point(size = 2.5, alpha = 0.8) +          # Make points clear but slightly transparent
      stat_ellipse(type = "t", level = 0.95, linetype = 2) +
      theme_bw() +                                   # Clean background theme
      theme(legend.position = "bottom")+              # Move legend to bottom so it doesn't crowd the margins
      scale_color_manual(name = "Cohort",
                     values = cohort_colors)+
      geom_label(x = box$box_x, y = box$box_y, label = box$box_label,
             fill = "white", color = "black", fontface = "bold", size = 4) +
      ggtitle(title_str)

  # Fixed axis limits suit the relative-abundance PCoA scale; CLR/PCA lives on a
  # much larger scale, so let ggplot auto-scale the axes in that case.
  if(norm_method != "clr" & !(norm_method == "identity" & ord_method == "euclidean")){
    base_plot <- base_plot +
      xlim(-.6, .6)+
      ylim(-.6, .6)
  }

  # Add the marginal distributions using ggExtra
final_plot <- ggMarginal(base_plot, 
                         type = "density",       # Change to "boxplot" or "histogram" if preferred
                         groupColour = TRUE, 
                         groupFill = TRUE, 
                         alpha = 0.4)            # Transparency for overlapping distributions


final_plot

  
}
```

```{r}
singleton_ps <- filter_taxa(ps, function(x) sum(x > 0) ==1, prune = TRUE)
singleton_ps
```

```{r}
filter_by_prevalence <- function(x, phylo_obj=ps){
prevalence_threshold <- x * nsamples(phylo_obj)

# Filter the taxa
filtered_ps <- filter_taxa(phylo_obj, function(x) sum(x > 0) >= prevalence_threshold, prune = TRUE)
filtered_ps <- prune_samples(sample_sums(filtered_ps) > 0, filtered_ps)

filtered_ps

}

filter_by_prevalence(.1)
filter_by_prevalence(.05)
```

```{r}
filtered_ps <- filter_by_prevalence(.05)
```

## Rarefaction

```{r}
source("~/Documents/ODSi/ODSiData/Functions/plot_rarefaction_curves.R")
```

```{r}
plot_rarefaction_curves(filtered_ps)
```

# Debias M

## 1) Confirm debias M is installed and environment is launched

```         
conda activate debiasm
```

## 2) Export data from R to python usable

```{r}
## Taxa rows × sample cols (same orientation required by ComBat-seq)
count_matrix <- as(otu_table(filtered_ps), "matrix")
if (!taxa_are_rows(filtered_ps)) {
  count_matrix <- t(count_matrix)
}

count_meta <- data.frame(sample_data(filtered_ps))
count_meta <- count_meta[colnames(count_matrix), , drop = FALSE]

batch_var <- droplevels(factor(count_meta$study))

## 1) Identify taxa non-zero in at least one sample of at least two studies
shared_taxa <- rowSums(sapply(levels(batch_var), function(b) {
  rowSums(count_matrix[, batch_var == b, drop = FALSE]) > 0
})) >= 2

## 2) Filter to shared taxa
shared_count_matrix <- count_matrix[shared_taxa, , drop = FALSE]

## 3) Drop samples with zero total counts after filtering
sample_keep <- colSums(shared_count_matrix) > 0
shared_count_matrix <- shared_count_matrix[, sample_keep, drop = FALSE]

## 4) Transpose to samples × taxa and write counts CSV
counts_df <- as.data.frame(t(shared_count_matrix))
counts_df <- cbind(sampleid = rownames(counts_df), counts_df)

write.csv(counts_df,
          "/Users/sophiehuebler/Documents/ODSi/ODSiData/Correcting/DebiasM/debiasm-counts.csv",
          row.names = FALSE)

## 5) Build metadata aligned to filtered sample order and write meta CSV
debiasm_meta <- count_meta[sample_keep, , drop = FALSE]
debiasm_meta <- debiasm_meta[, c("sample.id", "study", "agvhd")]
names(debiasm_meta)<- c("sampleid", "study", "agvhd")

write.csv(debiasm_meta,
          "/Users/sophiehuebler/Documents/ODSi/ODSiData/Correcting/DebiasM/debiasm-meta.csv",
          row.names = FALSE)
```

## 3) Run Debias-M python script

```         
conda activate debiasm
python /Users/sophiehuebler/Documents/ODSi/ODSiData/Correcting/DebiasM/run_debiasm.py
```

## 4) Import back into R

```{r}
corrected_df <- read.csv(
  "/Users/sophiehuebler/Documents/ODSi/ODSiData/Correcting/DebiasM/debiasm-corrected.csv",
  row.names = 1,
  check.names = FALSE
)

# corrected_df is samples × taxa; prune uncorrected_ps to the retained
# samples and taxa before swapping the otu_table
retained_samples <- rownames(corrected_df)
retained_taxa    <- colnames(corrected_df)

debiasm_ps <- prune_samples(sample_names(ps) %in% retained_samples,
                             ps)
debiasm_ps <- prune_taxa(taxa_names(debiasm_ps) %in% retained_taxa,
                          debiasm_ps)

# Align corrected matrix to the object's current sample and taxa order,
# then transpose to taxa × samples for otu_table assignment
corrected_mat <- as.matrix(corrected_df[sample_names(debiasm_ps),
                                         taxa_names(debiasm_ps)])
stopifnot(identical(colnames(corrected_mat), taxa_names(debiasm_ps)))
stopifnot(identical(rownames(corrected_mat), sample_names(debiasm_ps)))

otu_table(debiasm_ps) <- otu_table(t(corrected_mat), taxa_are_rows = TRUE)
```

```{r}
print("Matching otu rownames and tree tip labels:")
all(rownames(debiasm_ps@otu_table) %in% debiasm_ps@phy_tree$tip.label)

print(sprintf("OTU features: %d ; tree tips: %d", ntaxa(debiasm_ps@otu_table), length(debiasm_ps@phy_tree$tip.label)))
print(sprintf("Intersection: %d", sum(rownames(debiasm_ps@otu_table) %in% debiasm_ps@phy_tree$tip.label)))

print(sprintf("OTU sample: %d ; meta samples: %d", ncol(debiasm_ps@otu_table), nrow(debiasm_ps@sam_data)))
print(sprintf("Intersection: %d", sum(colnames(debiasm_ps@otu_table) %in% rownames(debiasm_ps@sam_data))))
```

```{r}
print(debiasm_ps)
```

## 5) Beta diversity

> These relative-abundance panels are most comparable to the logCPM uncorrected/ComBat panels, not the rarefied-to-900 ones.

```{r}
set.seed(9134)
```

### Bray

```{r}
uncorrected_beta_bray <- beta_pipe(filtered_ps,
          rare_depth = 1200,
          "bray",
"Beta Diversity (Bray-Curtis)
Rarefied 1200
Relative Abundance",
"rarefy",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

uncorrected_beta_bray
```

```{r}
debiasm_beta_bray <- beta_pipe(debiasm_ps,
          NULL,
          "bray",
"Beta Diversity (Bray-Curtis)
DEBIAS-M corrected
Relative Abundance",
"identity",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

debiasm_beta_bray
```

## Weighted Unifrac

```{r}
uncorrected_beta_wunifrac <- beta_pipe(filtered_ps,
          rare_depth = 1200,
          "wunifrac",
"Beta Diversity (Weighted Unifrac)
Rarefied 1200
Relative Abundance",
"rarefy",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

uncorrected_beta_wunifrac
```

```{r}
debiasm_beta_wunifrac <- beta_pipe(debiasm_ps,
          NULL,
          "wunifrac",
"Beta Diversity (Bray-Weighted Unifrac)
DEBIAS-M corrected
Relative Abundance",
"identity",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

debiasm_beta_wunifrac
```
