---
title: "Untitled"
format: html
editor: visual
---

# Libraries

```{r}
library(phyloseq)
library(qiime2R)
library(biomformat)
library(readr)
library(tidyverse)
library(ape)
library(Biostrings)
library(ggtree)
library(sva)
library(vegan)
library(ggExtra)
```

```{r}
source("~/Documents/ODSi/ODSiData/Functions/check_phyloseq.R")
source("~/Documents/ODSi/ODSiData/Functions/beta_pipe.R")
source("~/Documents/ODSi/ODSiData/Functions/parse_taxonomy.R")
source("~/Documents/ODSi/ODSiData/Functions/color_tips.R")
source("~/Documents/ODSi/ODSiData/Functions/filter_by_prevalence.R")
source("~/Documents/ODSi/ODSiData/Functions/plot_rarefaction_curves.R")
```

# Files

```{r}
table_qza <- "~/Documents/ODSi/ODSiData/Merging/full_cohort/merged-table.qza"
taxonomy_qza  <- "~/Documents/ODSi/ODSiData/Merging/full_cohort/taxonomy-gg.qza"
repseqs_qza   <- "~/Documents/ODSi/ODSiData/Merging/full_cohort/merged-seqs.qza"      
tree_nwk     <- "~/Documents/ODSi/ODSiData/Merging/full_cohort/merged-tree.nwk"
meta_tsv <- "~/Documents/ODSi/ODSiData/Merging/merged_metadata.tsv"
```

```{r}
qt_orig <- read_qza(table_qza)$data
tax_q  <- read_qza(taxonomy_qza)$data
tree    <- ape::read.tree(tree_nwk)
rep_seqs <- tryCatch(read_qza(repseqs_qza)$data, error = function(e) NULL)
meta <- read_tsv(meta_tsv)
```

# Phyloseq

## OTU

```{r}
otu_mat <- otu_table(qt_orig, taxa_are_rows = TRUE)
```

## Taxonomy

```{r}
tax_mat <- parse_taxonomy(tax_q)
tax_mat <- tax_table(tax_mat)
```

## Seqs

```{r}
seqs <- Biostrings::DNAStringSet(rep_seqs)
names(seqs) <- names(rep_seqs)
```

## Meta

```{r}

setdiff(meta$`sample-id`, colnames(otu_mat))->excluded
meta_df <- sample_data(meta %>%
                         filter(!(`sample-id` %in% excluded)))
rownames(meta_df)<- meta_df$`sample-id`
```

## Combining

```{r}
ps <- phyloseq(otu_mat, tax_mat, meta_df, phy_tree(tree), refseq(seqs))
print(ps)
```

```{r}
check_phyloseq(ps)
```

# Visualizing

```{r}
circular_tree <- ggtree(ps@phy_tree, layout = "circular", branch.length = "none") +
  geom_tippoint( size = 1.6, alpha = 0.9)

circular_tree
```

```{r}
genus_tree <- vizualize_collapse(ps, "Genus", show_labels = TRUE)
```

# Beta diversity

```{r}
singleton_ps <- filter_taxa(ps, function(x) sum(x > 0) ==1, prune = TRUE)
singleton_ps
```

```{r}


filter_by_prevalence(.1)
filter_by_prevalence(.05)
```

```{r}
filtered_ps <- filter_by_prevalence(.05)
```

## Rarefaction

```{r}
plot_rarefaction_curves(filtered_ps)
```

# Debias M

## 1) Confirm debias M is installed and environment is launched

```         
conda activate debiasm
```

## 2) Export data from R to python usable

```{r}
## Taxa rows × sample cols (same orientation required by ComBat-seq)
count_matrix <- as(otu_table(filtered_ps), "matrix")
if (!taxa_are_rows(filtered_ps)) {
  count_matrix <- t(count_matrix)
}

count_meta <- data.frame(sample_data(filtered_ps))
count_meta <- count_meta[colnames(count_matrix), , drop = FALSE]

batch_var <- droplevels(factor(count_meta$study))

## 1) Identify taxa non-zero in at least one sample of at least two studies
shared_taxa <- rowSums(sapply(levels(batch_var), function(b) {
  rowSums(count_matrix[, batch_var == b, drop = FALSE]) > 0
})) >= 2

## 2) Filter to shared taxa
shared_count_matrix <- count_matrix[shared_taxa, , drop = FALSE]

## 3) Drop samples with zero total counts after filtering
sample_keep <- colSums(shared_count_matrix) > 0
shared_count_matrix <- shared_count_matrix[, sample_keep, drop = FALSE]

## 4) Transpose to samples × taxa and write counts CSV
counts_df <- as.data.frame(t(shared_count_matrix))
counts_df <- cbind(sampleid = rownames(counts_df), counts_df)

write.csv(counts_df,
          "/Users/sophiehuebler/Documents/ODSi/ODSiData/Correcting/DebiasM/debiasm-counts.csv",
          row.names = FALSE)

## 5) Build metadata aligned to filtered sample order and write meta CSV
debiasm_meta <- count_meta[sample_keep, , drop = FALSE]
debiasm_meta <- debiasm_meta[, c("sample.id", "study", "agvhd")]
names(debiasm_meta)<- c("sampleid", "study", "agvhd")

write.csv(debiasm_meta,
          "/Users/sophiehuebler/Documents/ODSi/ODSiData/Correcting/DebiasM/debiasm-meta.csv",
          row.names = FALSE)
```

## 3) Run Debias-M python script

```         
conda activate debiasm
python /Users/sophiehuebler/Documents/ODSi/ODSiData/Correcting/DebiasM/run_debiasm.py
```

## 4) Import back into R

```{r}
corrected_df <- read.csv(
  "/Users/sophiehuebler/Documents/ODSi/ODSiData/Correcting/DebiasM/debiasm-corrected.csv",
  row.names = 1,
  check.names = FALSE
)

# corrected_df is samples × taxa; prune uncorrected_ps to the retained
# samples and taxa before swapping the otu_table
retained_samples <- rownames(corrected_df)
retained_taxa    <- colnames(corrected_df)

debiasm_ps <- prune_samples(sample_names(ps) %in% retained_samples,
                             ps)
debiasm_ps <- prune_taxa(taxa_names(debiasm_ps) %in% retained_taxa,
                          debiasm_ps)

# Align corrected matrix to the object's current sample and taxa order,
# then transpose to taxa × samples for otu_table assignment
corrected_mat <- as.matrix(corrected_df[sample_names(debiasm_ps),
                                         taxa_names(debiasm_ps)])
stopifnot(identical(colnames(corrected_mat), taxa_names(debiasm_ps)))
stopifnot(identical(rownames(corrected_mat), sample_names(debiasm_ps)))

otu_table(debiasm_ps) <- otu_table(t(corrected_mat), taxa_are_rows = TRUE)
```

```{r}
print("Matching otu rownames and tree tip labels:")
all(rownames(debiasm_ps@otu_table) %in% debiasm_ps@phy_tree$tip.label)

print(sprintf("OTU features: %d ; tree tips: %d", ntaxa(debiasm_ps@otu_table), length(debiasm_ps@phy_tree$tip.label)))
print(sprintf("Intersection: %d", sum(rownames(debiasm_ps@otu_table) %in% debiasm_ps@phy_tree$tip.label)))

print(sprintf("OTU sample: %d ; meta samples: %d", ncol(debiasm_ps@otu_table), nrow(debiasm_ps@sam_data)))
print(sprintf("Intersection: %d", sum(colnames(debiasm_ps@otu_table) %in% rownames(debiasm_ps@sam_data))))
```

```{r}
print(debiasm_ps)
```

## 5) Beta diversity

> These relative-abundance panels are most comparable to the logCPM uncorrected/ComBat panels, not the rarefied-to-900 ones.

```{r}
set.seed(9134)
```

### Bray

```{r}
uncorrected_beta_bray <- beta_pipe(filtered_ps,
          rare_depth = 1200,
          "bray",
"Beta Diversity (Bray-Curtis)
Rarefied 1200
Relative Abundance",
"rarefy",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

uncorrected_beta_bray
```

```{r}
debiasm_beta_bray <- beta_pipe(debiasm_ps,
          NULL,
          "bray",
"Beta Diversity (Bray-Curtis)
DEBIAS-M corrected
Relative Abundance",
"identity",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

debiasm_beta_bray
```

## Weighted Unifrac

```{r}
uncorrected_beta_wunifrac <- beta_pipe(filtered_ps,
          rare_depth = 1200,
          "wunifrac",
"Beta Diversity (Weighted Unifrac)
Rarefied 1200
Relative Abundance",
"rarefy",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

uncorrected_beta_wunifrac
```

```{r}
debiasm_beta_wunifrac <- beta_pipe(debiasm_ps,
          NULL,
          "wunifrac",
"Beta Diversity (Bray-Weighted Unifrac)
DEBIAS-M corrected
Relative Abundance",
"identity",
box_label_positions = c("l", "l"),
box_label_percs = c(.99, .89))

debiasm_beta_wunifrac
```
