---
title: "Untitled"
format: html
editor: visual
---

# Libraries

```{r}
library(phyloseq)
library(qiime2R)
library(biomformat)
library(readr)
library(tidyverse)
library(ape)
library(Biostrings)
library(ggtree)
library(sva)
library(vegan)
library(ggExtra)
```

```{r}
source("~/Documents/ODSi/ODSiData/Functions/check_phyloseq.R")
source("~/Documents/ODSi/ODSiData/Functions/beta_pipe.R")
source("~/Documents/ODSi/ODSiData/Functions/parse_taxonomy.R")
source("~/Documents/ODSi/ODSiData/Functions/color_tips.R")
source("~/Documents/ODSi/ODSiData/Functions/filter_by_prevalence.R")
source("~/Documents/ODSi/ODSiData/Functions/plot_rarefaction_curves.R")
source("~/Documents/ODSi/ODSiData/Functions/debiasm.R")
source("~/Documents/ODSi/ODSiData/Functions/build_phyloseq.R")
source("~/Documents/ODSi/ODSiData/Functions/plot_beta_correction.R")
source("~/Documents/ODSi/ODSiData/Functions/suppress_all.R")
```

# Build phyloseq

Building from the Merging/full_cohort folder

```{r}
ps <- build_phyloseq("~/Documents/ODSi/ODSiData/Merging/full_cohort",
                     prefix   = "merged-",
                     meta_tsv = "~/Documents/ODSi/ODSiData/Merging/merged_metadata.tsv")
```

```{r}
check_phyloseq(ps)
```

# Visualizing

```{r}
genus_tree <- vizualize_collapse(ps, "Genus", show_labels = TRUE)

genus_tree
```

# Metadata

```{r}
metadata <- ps@sam_data %>% as.data.frame()%>%
  unclass()%>%as.data.frame()
```

## Harmonization provenance

`Merging/Harmonization/harmonization_provenance.tsv` records, for every target column and every study, how that column was obtained from the study's parsed metadata. Statuses use the seven-level vocabulary of the `Harmonization_Summary` sheet in `data_dictionaries_harmonized.xlsx`, audited against `Merging/Harmonization/harmonize_gvhd_metadata.R`. Regenerate the file with `python3 Merging/Harmonization/build_provenance.py <out.tsv>` after changing the harmonizer.

```{r}
source("~/Documents/ODSi/ODSiData/Functions/study_colors.R")

status_levels <- c("exact", "direct", "recoded", "derived",
                   "limited", "unsupported", "unavailable")

status_key <- tibble::tribble(
  ~status,       ~definition,
  "exact",       "Copied verbatim from the source column; no modification.",
  "direct",      "Read off a single source column, type coercion only.",
  "recoded",     "Source values mapped or binned onto a canonical vocabulary.",
  "derived",     "Computed or generated from one or more source columns.",
  "limited",     "Populated, but with a documented restriction on coverage or meaning.",
  "unsupported", "The study records something related that cannot support this column.",
  "unavailable", "The study does not record it."
)

provenance <- read_tsv(
  "~/Documents/ODSi/ODSiData/Merging/Harmonization/harmonization_provenance.tsv",
  show_col_types = FALSE
) %>%
  mutate(
    study  = study_key(study),
    status = factor(status, levels = status_levels)
  )

stopifnot(!any(is.na(provenance$status)))

# Schema order and block order, both carried by the provenance file.
target_cols <- provenance %>%
  distinct(target_order, target_column) %>%
  arrange(target_order) %>%
  pull(target_column)

domain_levels <- provenance %>%
  distinct(domain_order, domain) %>%
  arrange(domain_order) %>%
  pull(domain)
```

## Missingness

Percent of each study's samples with no value, by target variable. A cross marks a variable the study does not record at all (`unavailable` or `unsupported`), as distinct from one it records but records incompletely.

```{r}
#| fig-width: 9
#| fig-height: 11

meta_flat <- metadata %>%
  rownames_to_column("sample-id") %>%
  mutate(.study = study_key(study))

missingness <- meta_flat %>%
  select(.study, any_of(target_cols)) %>%
  # phyloseq hands these back as characters, so empty strings and a literal
  # "NA" are missing values that is.na() alone would not catch.
  mutate(across(-.study, ~ na_if(na_if(str_squish(as.character(.x)), ""), "NA"))) %>%
  pivot_longer(-.study, names_to = "target_column", values_to = "value") %>%
  group_by(study = .study, target_column) %>%
  summarise(n = n(), n_missing = sum(is.na(value)), .groups = "drop") %>%
  mutate(pct_missing = 100 * n_missing / n)

heat <- missingness %>%
  left_join(provenance, by = c("study", "target_column")) %>%
  mutate(
    not_collected = status %in% c("unavailable", "unsupported"),
    domain        = factor(domain, levels = domain_levels),
    target_column = factor(target_column, levels = rev(target_cols)),
    study         = factor(study, levels = studies)
  )

# The provenance file and the data must agree: nothing the harmonizer populates
# may be empty everywhere, and nothing it calls absent may carry values.
stopifnot(
  !any(is.na(heat$status)),
  sum(heat$pct_missing == 100 & !heat$not_collected) == 0,
  sum(heat$pct_missing <  100 &  heat$not_collected) == 0
)

study_labs <- meta_flat %>%
  count(study = .study) %>%
  mutate(lab = paste0(study, "\n(n = ", n, ")")) %>%
  { setNames(.$lab, .$study) }

p_missing <- ggplot(heat, aes(x = study, y = target_column, fill = pct_missing)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_point(
    data  = function(d) dplyr::filter(d, not_collected),
    shape = 4, size = 1.4, stroke = 0.5, colour = "white", alpha = 0.85
  ) +
  facet_grid(domain ~ ., scales = "free_y", space = "free_y", switch = "y") +
  scale_fill_gradient(
    low = "#EAF2F8", high = "#0B3D66", limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100), labels = scales::label_number(suffix = "%"),
    name = "Missing"
  ) +
  scale_x_discrete(labels = study_labs, position = "top") +
  labs(
    x = NULL, y = NULL,
    title    = "Missingness in the harmonized metadata",
    subtitle = "Percent of each study's samples with no value, by target variable",
    caption  = str_wrap(paste(
      "A cross marks a variable the study does not record at all (harmonization status",
      "unavailable or unsupported), as distinct from one it records but records incompletely.",
      "Statuses are audited from Merging/Harmonization/harmonize_gvhd_metadata.R."
    ), width = 110)
  ) +
  theme_minimal(base_size = 9) +
  theme(
    panel.grid            = element_blank(),
    axis.text.y           = element_text(family = "mono", size = 6),
    axis.text.x           = element_text(size = 7.5, lineheight = 1.1),
    strip.text.y.left     = element_text(angle = 0, hjust = 0, size = 7.5,
                                         colour = "grey25"),
    strip.placement       = "outside",
    panel.spacing.y       = unit(2, "pt"),
    legend.position       = "right",
    legend.key.width      = unit(8, "pt"),
    plot.title            = element_text(face = "bold", size = 11),
    plot.subtitle         = element_text(colour = "grey30", size = 8,
                                         margin = margin(b = 6)),
    plot.caption          = element_text(hjust = 0, colour = "grey35", size = 6.5,
                                         lineheight = 1.2),
    plot.title.position   = "plot",
    plot.caption.position = "plot"
  )

p_missing
```

```{r}
ggsave("~/Documents/ODSi/ODSiData/Merging/metadata_missingness.png",
       p_missing, width = 9, height = 11, dpi = 300, bg = "white")
```

## Provenance table

How each target variable was obtained, by study. The fill groups the seven statuses into the four things that can happen to a variable: taken as given (blue), collapsed onto a canonical vocabulary (green), inferred from other columns (orange), or not obtainable (grey). The status word carries the identity, so the fill is secondary and the table reads without color.

```{r}
provenance_wide <- provenance %>%
  select(domain_order, domain, target_order, target_column, study, status) %>%
  mutate(study = factor(study, levels = studies)) %>%
  pivot_wider(names_from = study, values_from = status) %>%
  arrange(domain_order, target_order) %>%
  select(-domain_order, -target_order)

status_fill <- c(
  exact       = "#D7E6F4",  # taken as given
  direct      = "#E7F0F9",
  recoded     = "#D9EBDD",  # collapsed onto a canonical vocabulary
  derived     = "#FAE7CD",  # inferred from other columns
  limited     = "#F6D7C4",
  unsupported = "#E4E4E4",  # not obtainable
  unavailable = "#F6F6F6"
)

prov_tbl <- provenance_wide %>%
  gt::gt(groupname_col = "domain", rowname_col = "target_column") %>%
  gt::tab_header(
    title    = "Harmonization provenance by variable and study",
    subtitle = "How each target column was obtained from each study's parsed metadata"
  ) %>%
  gt::tab_source_note(gt::md(
    paste0("**", status_key$status, "** &mdash; ", status_key$definition,
           collapse = "<br>")
  )) %>%
  gt::tab_style(
    style     = gt::cell_text(font = "IBM Plex Mono, Menlo, monospace"),
    locations = gt::cells_stub()
  ) %>%
  gt::tab_options(
    row_group.font.weight = "bold",
    data_row.padding      = gt::px(3),
    table.font.size       = gt::px(11)
  )

# One fill per study column per status. gt takes row indices here, so the
# lookup is done in plain R rather than as a tidy expression.
for (st in studies) {
  for (s in status_levels) {
    idx <- which(as.character(provenance_wide[[st]]) == s)
    if (length(idx) == 0L) next
    prov_tbl <- prov_tbl %>%
      gt::tab_style(
        style     = gt::cell_fill(color = unname(status_fill[[s]])),
        locations = gt::cells_body(columns = all_of(st), rows = idx)
      )
  }
}

prov_tbl
```

```{r}
#| eval: false
# Long form, for the rule and caveat behind any single cell.
provenance %>%
  filter(target_column == "gut_gvhd") %>%
  select(study, status, source_column, harmonization_rule) %>%
  arrange(study)
```

# Beta diversity

## How many taxa appear in only 1 study?

```{r}
singleton_ps <- filter_taxa(ps, function(x) sum(x > 0) ==1, prune = TRUE)
singleton_ps
```

## Filter by prevelance

```{r}
filtered_ps <- filter_by_prevalence(.1)
filtered_ps
```

## Rarefaction

```{r}
plot_rarefaction_curves(filtered_ps, aggregate = "mean")
```

# Plotting before/after correction

```{r}
debiasm_ps <- debiasm_correct(filtered_ps)
```

```{r}
check_phyloseq(debiasm_ps)
```

## Full cohort

### Bray

```{r}
plot_beta_correction("!is.na(study)", 
                     ord_method = "bray",
                     phylo_obj = filtered_ps,
                     rare_depth = 900)
```

### Weighted Unifrac

```{r}
plot_beta_correction("!is.na(study)", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900)
```

## Exclude Ingham

```{r}

without_ingham <- plot_beta_correction("study != 'Ingham'", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900)


```

```{r}
print(without_ingham)
```

```{r}
baseline_without_ingham <- 
  plot_beta_correction("study != 'Ingham' & timepoint %in% c('conditioning', 'transplant')", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900)
```

```{r}
before_engraftment<- 
  plot_beta_correction("timepoint!='follow-up'", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900)
```

# Baseline

```{r}
baseline_ps <- subset_ps_func(filtered_ps,
                            "timepoint %in% c('pre-conditioning','conditioning', 'transplant') & disease != 'none'")
print(baseline_ps)
```

```{r}
baseline<- plot_beta_correction("!is.na(study)", 
                     ord_method = "wunifrac",
                     phylo_obj = baseline_ps,
                     rare_depth = 900)
```

```{r}
baseline_adults_viz <- plot_beta_correction("!is.na(study) & age >= 17 & study != 'DAmico'", 
                     ord_method = "wunifrac",
                     phylo_obj = baseline_ps,
                     rare_depth = 900)
```

```{r}
baseline_adults <- plot_beta_correction("!is.na(study) & age >= 17", "wunifrac", baseline_ps, label_corner = "br") 
```

```{r}
baseline_adults_ps <- subset_ps_func(baseline_ps,
                            "!is.na(study) & age >= 17")

baseline_adults_genus <- vizualize_collapse_dots(baseline_adults_ps, "Genus",
                        id_col_name = "sample.id", cohort_col_name = "study",
                        show_labels = FALSE, dot_spacing = 0.8)

baseline_adults_species <- vizualize_collapse_dots(baseline_adults_ps, "Species",
                        id_col_name = "sample.id", cohort_col_name = "study",
                        show_labels = FALSE, dot_spacing = 0.8)
```

```{r}
baseline_children <- plot_beta_correction("study != 'Vallet' & age < 17", 
                     ord_method = "wunifrac",
                     phylo_obj = baseline_ps,
                     rare_depth = 900)
```

# No stratification

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none'", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none'", 
                     ord_method = "bray",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```

# Stratified by Age

## Adults (N = 662)

```{r}

plot_beta_correction("!is.na(disease)& disease != 'none' & study %in% c('Artacho', 'Fujimoto', 'Liu', 'Vallet')", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )

```

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none' & study %in% c('Artacho', 'Fujimoto', 'Liu', 'Vallet')", 
                     ord_method = "bray",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```

## Children (N = 199)

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none' & !(study %in% c('Artacho', 'Fujimoto', 'Liu', 'Vallet'))", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none' & !(study %in% c('Artacho', 'Fujimoto', 'Liu', 'Vallet'))", 
                     ord_method = "bray",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```

# Stratified by before/after engraftment

## Before engraftment (N = 442)

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none' & timepoint %in% c('pre-conditioning', 'conditioning', 'transplant', 'pre-engraftment')", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none' & !(timepoint %in% c('pre-conditioning', 'conditioning', 'transplant', 'pre-engraftment'))", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```

# Stratified by age and timepoint

-   Adults: Artacho, Fujimoto, Jarosch, Liu, Vallet

-   Children: Ingham, Damico

```{r}
filtered_metadata <- filtered_ps@sam_data %>% as.data.frame()%>%
  unclass()%>%as.data.frame()
```

```{r}
sample_categories <- filtered_metadata %>%
  mutate(is_member = ifelse(!is.na(disease) & disease == "none", FALSE, TRUE))%>%
  mutate(Age_group = ifelse(study %in% c("Artacho", "Fujimoto", 
                                      "Jarosch", "Liu", "Vallet"),
                         "Adults", "Children"))%>%
  select(sample.id, Age_group, timepoint, is_member)%>%
  pivot_wider(id_cols = sample.id,
              names_from = c(Age_group, timepoint),
              values_from = is_member,
              values_fill = FALSE,
              names_glue = "{Age_group}_{timepoint}")

names(sample_categories)
```

```{r}
adults_preconditioning <- prune_samples(sample_categories$`Adults_pre-conditioning`, filtered_ps)
debiasm_adults_preconditioning <- debiasm_correct(adults_preconditioning)

adults_conditioning <- prune_samples(sample_categories$`Adults_conditioning`, filtered_ps)

adults_transplant<- prune_samples(sample_categories$`Adults_transplant`, filtered_ps)

adults_preengraftment <- prune_samples(sample_categories$`Adults_pre-engraftment`, filtered_ps)

adults_engraftment <- prune_samples(sample_categories$`Adults_engraftment`, filtered_ps)

adults_followup <- prune_samples(sample_categories$`Adults_follow-up`, filtered_ps)



```

```{r}
children_preconditioning <- prune_samples(sample_categories$`Children_pre-conditioning`, filtered_ps)

children_conditioning <- prune_samples(sample_categories$`Children_conditioning`, filtered_ps)

children_transplant<- prune_samples(sample_categories$`Children_transplant`, filtered_ps)

children_preengraftment <- prune_samples(sample_categories$`Children_pre-engraftment`, filtered_ps)

children_engraftment <- prune_samples(sample_categories$`Children_engraftment`, filtered_ps)

children_followup <- prune_samples(sample_categories$`Children_follow-up`, filtered_ps)
```

## Adults

### Preconditioning (N-112)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = adults_preconditioning,
                     rare_depth = 900
                     )

```

### Conditioning (N=86)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = adults_conditioning,
                     rare_depth = 900
                     )
```

### Transplant (N=78)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = adults_transplant,
                     rare_depth = 900
                     )
```

### Pre-engraftment (N=92)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = adults_preengraftment,
                     rare_depth = 900
                     )
```

### Engraftment (N=103)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = adults_engraftment,
                     rare_depth = 900
                     )
```

### Follow-up (N = 191)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = adults_followup,
                     rare_depth = 900
                     )
```

## Children

### Preconditioning (N=24)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = children_preconditioning,
                     rare_depth = 900
                     )

```

### Conditioning (N = 13)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = children_conditioning,
                     rare_depth = 900
                     )
```

### Transplant (N=11)

```{r}
# plot_beta_correction("study != 'random'", 
#                      ord_method = "wunifrac",
#                      phylo_obj = children_transplant,
#                      rare_depth = 900
#                      )
```

### Pre-engraftment (N = 26)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = children_preengraftment,
                     rare_depth = 900
                     )
```

### Engraftment (N = 24)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = children_engraftment,
                     rare_depth = 900
                     )
```

### Follow-up (N = 101)

```{r}
plot_beta_correction("study != 'random'", 
                     ord_method = "wunifrac",
                     phylo_obj = children_followup,
                     rare_depth = 900
                     )
```

# Stratified by age and engraftment

## Adults

### Before (N = 368)

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none' & timepoint %in% c('pre-conditioning', 'conditioning', 'transplant', 'pre-engraftment') & study %in% c('Artacho', 'Fujimoto', 'Liu', 'Vallet')", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```

### After (N = 294)

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none' & !(timepoint %in% c('pre-conditioning', 'conditioning', 'transplant', 'pre-engraftment')) & study %in% c('Artacho', 'Fujimoto', 'Liu', 'Vallet')", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```

## Children

### Before (N = 74)

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none' & timepoint %in% c('pre-conditioning', 'conditioning', 'transplant', 'pre-engraftment') & !(study %in% c('Artacho', 'Fujimoto', 'Liu', 'Vallet'))", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```

### After (N = 125)

```{r}
plot_beta_correction("!is.na(disease)& disease != 'none' & !(timepoint %in% c('pre-conditioning', 'conditioning', 'transplant', 'pre-engraftment')) & !(study %in% c('Artacho', 'Fujimoto', 'Liu', 'Vallet'))", 
                     ord_method = "wunifrac",
                     phylo_obj = filtered_ps,
                     rare_depth = 900
                     )
```
