---
title: "Read Attrition"
format: html
editor: visual
---

```{r}
#| include: false

library(tidyverse)
library(qiime2R)
library(ggh4x)
library(scales)
```

# Goal

Examine read attrition across microbiome processing pipeline

# Steps

1.  Original read depth

    -   Per-sample sequence count in `QiimeData/demux.qza` as written by `02_import.slurm` / `02_import_single.slurm` — the demultiplexed ENA/SRA FASTQs before any trimming or denoising.
    -   For paired-end studies this counts read *pairs* (one count per fragment, not per mate), so it stays on the same scale as every downstream step.

2.  After trimming primers

    -   Paired-end studies (`03_trim_paired.slurm`): `qiime cutadapt trim-paired` removes `PRIMER_F` / `PRIMER_R` as 5' front adapters, one per mate, with `--p-discard-untrimmed` — a pair in which either primer is not found is dropped.
    -   Single-end / pre-merged studies (`03_trim_single.slurm`): **two** sequential `trim-single` passes — pass 1 strips the 5' forward primer (`--p-discard-untrimmed`), pass 2 strips the 3' reverse-complemented reverse primer (`--p-no-discard-untrimmed`). Two passes are required because cutadapt's `--times 1` default removes at most one adapter per read; only pass 1 discards reads.
    -   Ingham2019 is a no-op here — its primers were already stripped upstream by `lib/prep_ingham_454.sh`, so `FWD_PRIMER` is empty and the raw demux carries through unchanged.

3.  Denoising (one step if single end, or as 3 steps with denoising and then merging and non-chimeric if paired end)

    -   Every DADA2 arm first applies an internal quality filter (the `filtered` column of `stats.qza`: truncation at `trunc-len`, an expected-error ceiling `max-ee`, and removal of any read containing an N). This sits between steps 2 and 3 and can optionally be plotted — see `step_colors_optional` below.
    -   **DADA2 `denoise-paired`** (Artacho2024, Fujimoto2024, Vallet2023): reads are truncated at `TRUNC_LEN_F/R`, an Illumina error model is learned per run, and exact amplicon sequence variants are inferred separately for each mate (**Denoised**); mates are then overlapped into a single amplicon and pairs that fail to overlap are dropped (**Merged**); finally `consensus` chimera detection removes bimeras (**Non-chimeric**).
    -   **DADA2 `denoise-single`** (Liu2017, DAmico2019): the same Illumina error model applied to un-merged forward reads (Liu2017) or to depositor-joined amplicons as a sensitivity arm (DAmico2019) — ASV inference (**Denoised**) then chimera removal (**Non-chimeric**). There is no merge column, so **Merged** is absent for these studies.
    -   **DADA2 `denoise-pyro`** (Ingham2019, Jarosch2023): the 454/Ion Torrent error model, which tolerates homopolymer indels and variable read length instead of truncating every read to a fixed length. Reports the same two substeps as `denoise-single` (**Denoised**, **Non-chimeric**).
    -   **Deblur `denoise-16S`** (alternate arm for pre-merged reads): a mandatory `quality-filter q-score` pre-step, then sub-OTU denoising against the positive 16S reference at a fixed `--p-trim-length`, which additionally discards every read shorter than that length. Deblur reports its own columns (reads-raw → artifact-filtered → hit-reference) rather than a chimera column, so its substeps are mapped onto **Denoised** / **Non-chimeric** when plotted.

4.  Quality control (qc step)

    -   `05_qc.slurm` runs `qiime feature-classifier classify-consensus-vsearch` on the study's rep-seqs against the Greengenes2 backbone at `perc-identity = 0.97`, `query-cov = 0.90`, `min-consensus = 0.51`.
    -   `qiime taxa filter-table` / `filter-seqs` then drop every feature classified `Unassigned`; the reads lost here are those carried by ASVs with no high-confidence taxonomic assignment.

5.  Closed reference ASV mapping (map step)

    -   `06_gg2_map.slurm` runs `qiime greengenes2 non-v4-16s` at 99% identity, re-expressing the QC-passing ASVs in the Greengenes2 backbone namespace; ASVs with no backbone match drop out along with their reads.
    -   This is run uniformly on every study — including V4-only Liu2017 — so that all primer regions end up in one namespace and align to the shared GG2 phylogeny for merging and UniFrac.

# Plotting

## Colors

### Study colors (facet strips)

```{r}
source("~/Documents/ODSi/ODSiData/Functions/study_colors.R")

# Facet strips: outline in the study's tree color, translucent fill behind the label.
# Order must match the factor level order of df_long$study.
base_study_colors        <- unname(study_colors[studies])
transparent_study_colors <- alpha(base_study_colors, 0.65)
```

### Pipeline step colors

```{r}
# Attrition palette, ordered along the pipeline:
#   Input          dark purple            raw demultiplexed depth
#   Trimmed        blue                   after cutadapt primer removal
#   Denoised   \
#   Merged      >  green -> yellow ramp   the DADA2 denoising substeps
#   Non-chimeric/                         (single-end arms simply omit "Merged")
#   QC             orange                 vsearch high-confidence assignment gate
#   Mapped         red                    Greengenes2 closed-reference mapping

# The denoising substeps are generated as a gradient so the block reads as one
# family; change the stops / n here if the substep count ever changes.
denoise_gradient <- colorRampPalette(c("#1B7A3F", "#DCE04A"))(3)
# -> "#1B7A3F" "#7BAD45" "#DCE04A"

step_colors <- c(
  "Input"        = "#3F2B56",           # darkest, purple
  "Trimmed"      = "#3C6FA8",           # blue
  "Denoised"     = denoise_gradient[1], # green
  "Merged"       = denoise_gradient[2], # green-yellow
  "Non-chimeric" = denoise_gradient[3], # yellow
  "QC"           = "#EF8A17",           # orange
  "Mapped"       = "#C1272D"            # red
)

# Draw order: largest (earliest) bar in back, smallest (latest) in front.
step_order <- names(step_colors)

# Optional: DADA2's internal quality-filter column ("filtered" in stats.qza),
# which sits between Trimmed and Denoised. Kept out of the default palette so the
# legend stays at seven entries; splice it in if you decide to plot it.
step_colors_optional <- c("Filtered" = "#7FA9D4")   # lighter shade of the trim blue
# step_colors <- append(step_colors, step_colors_optional, after = 2)
# step_order  <- names(step_colors)
```

```{r}
#| fig-height: 1.4
#| fig-width: 8

# Quick visual check of the ramp
tibble(step = factor(step_order, levels = step_order), y = 1) %>%
  ggplot(aes(x = step, y = y, fill = step)) +
  geom_col(width = 1) +
  scale_fill_manual(values = step_colors) +
  theme_void() +
  theme(axis.text.x = element_text(size = 8), legend.position = "none")
```

## Metadata

```{r}
meta <- read_tsv("/Users/sophiehuebler/Documents/ODSi/ODSiData/Merging/merged_metadata.tsv")
```

## Data frame

Read counts are extracted from the QIIME2 artifacts on CHPC by `chpc/lib/extract_read_counts.py` and committed to `Merging/ReadCounts/`, one wide TSV per study. Columns are the pipeline stages in order; stages a study's denoiser arm never produces (`trimmed` for Ingham2019, `merged` for the single-end / pyro arms) are carried forward from the previous stage so the bar draws flat rather than dropping to zero, and `stages_imputed` records which columns that applied to. `*_read_counts_raw.tsv` holds the same values pre-fill, with genuine gaps left empty.

```{r}
read_counts_dir <- "/Users/sophiehuebler/Documents/ODSi/ODSiData/Merging/ReadCounts"

# "_read_counts.tsv$" deliberately excludes the "_read_counts_raw.tsv" files.
count_files <- list.files(read_counts_dir,
                          pattern    = "_read_counts\\.tsv$",
                          full.names = TRUE)

read_counts <- count_files %>%
  map_dfr(read_tsv,
          col_types = cols(
            `sample-id`    = col_character(),
            study          = col_character(),
            stages_imputed = col_character(),
            .default       = col_double()
          )) %>%
  dplyr::rename(sample.id = `sample-id`) %>%
  # The CHPC read-count TSVs still label studies "Artacho2024"/"Liu2017";
  # the shared palette and the harmonized metadata both key on the bare
  # surname, so normalize here. study_key() errors on an unknown cohort.
  mutate(study = study_key(study))

# --- Checks -----------------------------------------------------------------
# 1. one row per sample
stopifnot(!any(duplicated(read_counts$sample.id)))
# 2. every study is in the shared palette, or the facet strip colors misalign
stopifnot(all(read_counts$study %in% studies))
# 3. read counts and metadata cover the same samples
stopifnot(setequal(read_counts$sample.id, meta$`sample-id`))

read_counts %>%
  count(study) %>%
  left_join(read_counts %>%
              group_by(study) %>%
              summarise(across(c(input, mapped), sum), .groups = "drop"),
            by = "study") %>%
  mutate(retained = scales::percent(mapped / input, accuracy = 0.1))
```

```{r}
# TSV column -> legend label. "filtered" is DADA2's internal quality-filter
# column; it is carried through here but only plotted if "Filtered" was spliced
# into step_colors above, since the filter below keys off step_order.
step_lookup <- c(
  input        = "Input",
  trimmed      = "Trimmed",
  filtered     = "Filtered",
  denoised     = "Denoised",
  merged       = "Merged",
  non_chimeric = "Non-chimeric",
  qc           = "QC",
  mapped       = "Mapped"
)

df_long <- read_counts %>%
  # Facet order must match the level order of base_study_colors.
  mutate(study = factor(study, levels = studies)) %>%
  # Sort samples by starting depth so each panel reads as a wedge.
  # Swap desc(input) for desc(mapped) to sort by final depth instead.
  arrange(study, desc(input)) %>%
  mutate(sample.id = factor(sample.id, levels = rev(unique(sample.id)))) %>%
  pivot_longer(
    cols      = all_of(names(step_lookup)),
    names_to  = "step_col",
    values_to = "Reads"
  ) %>%
  mutate(Pipeline_Step = factor(step_lookup[step_col], levels = step_order)) %>%
  # Drops any stage not in the active palette (i.e. "Filtered" by default).
  filter(!is.na(Pipeline_Step), !is.na(Reads)) %>%
  # Bars are overlaid with position = "identity", so row order sets draw order:
  # Input first (largest, in back) through Mapped last (smallest, in front).
  arrange(sample.id, Pipeline_Step) %>%
  select(sample.id, study, Pipeline_Step, Reads, stages_imputed)

df_long
```

## Plot

```{r}
plot_attrition_func <- function(df, log_reads = FALSE){
  if(log_reads){
    df$Reads <- log(df$Reads)
  }
  
  ggplot(df, aes(y = sample.id, x = Reads, fill = Pipeline_Step)) +
  geom_bar(stat = "identity", position = "identity") +
  scale_fill_manual(values = step_colors, breaks = step_order) +

  # Swap facet_grid for facet_grid2 from the ggh4x package
  facet_grid2(
    study ~ .,
    scales = "free_y",
    space = "free_y",
    switch = "y",

    strip = strip_themed(
      background_y = elem_list_rect(fill = transparent_study_colors, color = base_study_colors)
    )
  ) +

  labs(
    title = "Read Attrition Across Microbiome Processing Pipeline",
    y = "Individual Samples",
    x = "Number of Reads",
    fill = "Processing Step"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    #strip.placement = "outside",


    strip.text.y.left = element_text(size = 12, face = "bold", color = "black", angle = 0),
    panel.grid.major.y = element_blank(),
    legend.position = "bottom"
  ) +


  guides(fill = guide_legend(nrow = 1, byrow = TRUE))
}
```

```{r}
plot_attrition_func(df_long)
```

```{r}
plot_attrition_func(df_long, log_reads = TRUE)
```

# Options

```{r}
plot_attrition_func <- function(df, log_reads = FALSE, flip_vertical = FALSE, 
                                organize_stage = "import") {
  
  # 1. Organize samples based on the specified stage
  # Filters the dataframe to the target stage, sorts by Reads, and re-levels the sample.id factor
  if (organize_stage %in% unique(df$Pipeline_Step)) {
    stage_df <- df[df$Pipeline_Step == organize_stage, ]
    
    # Sort samples by reads. 
    # Note: decreasing = FALSE is used so that the sample with the most reads appears 
    # at the top of a horizontal plot or the right of a vertical plot.
    ordered_samples <- stage_df$sample.id[order(stage_df$Reads, decreasing = FALSE)]
    df$sample.id <- factor(df$sample.id, levels = ordered_samples)
  } else {
    warning(paste("Stage", organize_stage, "not found in Pipeline_Step. Skipping reordering."))
  }
  
  # 2. Handle log transformation and dynamic axis labeling
  if (log_reads) {
    df$Reads <- log(df$Reads)
    read_label <- "Log (Number of Reads)"
  } else {
    read_label <- "Number of Reads"
  }
  
  # 3. Build the plot based on orientation
  if (flip_vertical) {
    
    # --- VERTICAL ORIENTATION ---
    p <- ggplot(df, aes(x = sample.id, y = Reads, fill = Pipeline_Step)) +
      geom_bar(stat = "identity", position = "identity") +
      scale_fill_manual(values = step_colors, breaks = step_order) +
      
      # Adjusted facet_grid2 for vertical layout
      facet_grid2(
        . ~ study,
        scales = "free_x",
        space = "free_x",
        switch = "x",
        strip = strip_themed(
          background_x = elem_list_rect(fill = transparent_study_colors, color = base_study_colors)
        )
      ) +
      labs(
        title = "Read Attrition Across Microbiome Processing Pipeline",
        x = "Individual Samples",
        y = read_label, # Reads on the Y-axis
        fill = "Processing Step"
      ) +
      theme_minimal() +
      theme(
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text.x.bottom = element_text(size = 12, face = "bold", color = "black", angle = 0),
        panel.grid.major.x = element_blank(),
        legend.position = "bottom"
      )
    
  } else {
    
    # --- HORIZONTAL ORIENTATION (Original) ---
    p <- ggplot(df, aes(y = sample.id, x = Reads, fill = Pipeline_Step)) +
      geom_bar(stat = "identity", position = "identity") +
      scale_fill_manual(values = step_colors, breaks = step_order) +
      
      # Original facet_grid2 for horizontal layout
      facet_grid2(
        study ~ .,
        scales = "free_y",
        space = "free_y",
        switch = "y",
        strip = strip_themed(
          background_y = elem_list_rect(fill = transparent_study_colors, color = base_study_colors)
        )
      ) +
      labs(
        title = "Read Attrition Across Microbiome Processing Pipeline",
        y = "Individual Samples",
        x = read_label, # Reads on the X-axis
        fill = "Processing Step"
      ) +
      theme_minimal() +
      theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.text.y.left = element_text(size = 12, face = "bold", color = "black", angle = 0),
        panel.grid.major.y = element_blank(),
        legend.position = "bottom"
      )
  }
  
  # 4. Add common legend formatting and return
  p <- p + guides(fill = guide_legend(nrow = 1, byrow = TRUE))
  
  return(p)
}
```

```{r}
plot_attrition_func(df_long, flip_vertical = TRUE)
```
