---
title: "Parsing Fujimoto Metadata"
format: html
editor: visual
---

# Coding Set Up

```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(lubridate)
```

# Reading

The clinical data comes from supplementary table 2 (detailed patient characteristics) from the publication.

```{r}
fuji_samples_raw <- read.csv("~/Documents/ODSi/ODSiData/Fujimoto2024/Metadata/Raw_metadata/meta_samples.csv",
                     stringsAsFactors = FALSE, check.names = FALSE)
```

# Cleaning Samples

Filter for only the patient samples

```{r}
fuji_samples_human <- fuji_samples_raw %>%
  filter(Organism == "human metagenome")
```

Getting the constant columns

```{r}
constant_columns <- sapply(fuji_samples_human,
                           function(x) length(unique(x))==1)
paste(names(fuji_samples_human[constant_columns]), collapse = ", ")
```

```{r}
fuji_PCRinfo <- fuji_samples_human %>%
  select(`Assay Type`, 
         BioProject, BioSampleModel, `Center Name`, Collection_Date, Consent,
         geo_loc_name_country, geo_loc_name_country_continent, geo_loc_name, HOST,
         Instrument, lat_lon, LibraryLayout, LibrarySelection, LibrarySource,
         Organism, Platform, ReleaseDate, version, `SRA Study`,
         AvgSpotLen, Bases, Bytes, `DATASTORE provider`, `DATASTORE filetype`,
         `DATASTORE region`, create_date)

fuji_identifiers <- fuji_samples_human %>%
  select(BioSample, Experiment, `Library Name`, `isolation_source`)

fuji_samples_patients <- fuji_samples_human %>%
  select(-c(`Assay Type`, 
         BioProject, BioSampleModel, `Center Name`, Collection_Date, Consent,
         geo_loc_name_country, geo_loc_name_country_continent, geo_loc_name, HOST,
         Instrument, lat_lon, LibraryLayout, LibrarySelection, LibrarySource,
         Organism, Platform, ReleaseDate, version, `SRA Study`,
         AvgSpotLen, Bases, Bytes, `DATASTORE provider`, `DATASTORE filetype`,
         `DATASTORE region`, create_date, BioSample, Experiment, `Library Name`,
         `isolation_source`))
```

```{r}
fuji_samples <- fuji_samples_patients %>%
mutate(sample_name = str_remove( `Sample Name`, "OCUBMT"))%>%
  separate_wider_delim(
    cols = sample_name, 
    delim = "_", 
    names = c("Patient", "Timepoint")
  )%>%
  mutate(Timepoint = str_remove(Timepoint, "-.*"))%>%
  # mutate(Timepoint = case_when(Day == "pre" ~ NA,
  #                              TRUE ~ Day))%>%
  #mutate(Timepoint = as.numeric(Timepoint))%>%
  mutate(Baseline1 = ifelse(Timepoint == "pre", 1, 0),
         Baseline2 = ifelse(Timepoint == "day0", 1, 0))%>%
  mutate(id = Run)%>%
  select(
    id,
    Run,
    `Sample Name`,
    Patient,
    Timepoint,
    Baseline1,
    Baseline2
  )%>%
  arrange(Patient, Timepoint)
```

All patients have a preconditioning baseline (within 14days before the initial date of conditioning therapy) and 38 have a day 0 measurement.

```{r}
fuji_samples %>%
  group_by(Patient)%>%
  summarize(Has_Pre = "pre" %in% Timepoint,
            Has_Zero = "day0" %in% Timepoint)%>%
  filter(!Has_Pre | !Has_Zero)
```

# Cleaning Clinical

```{r}
fuji_clinical_raw <- readxl::read_excel("~/Documents/ODSi/ODSiData/Fujimoto2024/Metadata/Raw_metadata/meta_clinical.xlsx", col_types = "text")

fuji_clinical <- fuji_clinical_raw
```

# Linking samples and clinical

```{r}
fuji_meta_qiime <- fuji_samples %>%
  left_join(fuji_clinical %>%
              select(Patient, 
                     Age, Sex,
                     Disease,
                     Donor, Graft_Type,
                     AGVHD),
            by = "Patient")

names(fuji_meta_qiime)[1]<- "sample-id"
```

# Exporting

```{r}
write_tsv(fuji_meta_qiime, file = "/Users/sophiehuebler/Documents/ODSi/ODSiData/Fujimoto2024/Metadata/fuji_meta_qiime.tsv")
```

# Linking post filtering

One of the samples got removed in the importing to qiime or dada processing. So need to remove from the metadata.

```{r}
# original_meta <- read_tsv("/Users/sophiehuebler/Documents/ODSi/ODSiData/Fujimoto2024/Metadata/fuji_meta_qiime.tsv")
# 
# 
# filtered_meta <- read_tsv("/Users/sophiehuebler/Documents/ODSi/ODSiData/Fujimoto2024/Metadata/post_filtering.tsv", col_names = T)%>%
#   filter(`sample-id` != "#q2:types") %>%
#   type.convert(as.is = TRUE)
```

```{r}
#setdiff(original_meta$id, filtered_meta$`sample-id`)
```

```{r}
# fixed_meta <- original_meta %>%
#   filter(id != "SRR28984726")
# 
# names(fixed_meta)[1]<- "sample-id"
```

```{r}
#write_tsv(fixed_meta, file = "/Users/sophiehuebler/Documents/ODSi/ODSiData/Fujimoto2024/Metadata/fuji_meta_qiime.tsv")
```
