---
title: "Parsing Vallet2023 Metadata"
format: html
editor: visual
---

# Coding Set Up

```{r}
library(tidyverse)
library(janitor)
library(readxl)
library(lubridate)
```

## Parsing dates

```{r}
parse_mixed_dates <- function(x,
                              corrupted = F) {
  # Identify which values are purely numeric (Excel dates converted to strings)
  is_excel_num <- grepl("^[0-9]+$", x)
  
  # Step A: Convert the Excel numbers
  # We convert to numeric first, and handle the NAs
  num_dates <- excel_numeric_to_date(as.numeric(ifelse(is_excel_num, x, NA)))
  
  #if corrupted, then switch the day and month
  if (corrupted == TRUE) {
    
    num_dates <- make_date(
      year  = year(num_dates),
      month = day(num_dates),   # Swap day into month
      day   = month(num_dates)  # Swap month into day
    )
  }
  
  # Step B: Convert the text strings (like "26/05/2014")
  str_dates <- as.Date(parse_date_time(ifelse(!is_excel_num, x, NA), orders = "dmy"))
  
  # Step C: Combine them seamlessly
  coalesce(num_dates, str_dates)
  

}
```

# Vallet2023

## Reading

Samples has 118 observations belonging to 50 different people. Clinical has 154 observations belonging to 55 people. We have at most 50 people. We need to first link allozithro_id from the clinical data frame to the sample_name prefix in the samples data frame, both of which link to the individual person the observation belongs to. Additionally, the sample_name suffix in the samples data frame is a date which we can link to sample.time in the clinical data frame to determine which observation is baseline (closest to dat.hsct).

```{r}
vallet_samples_raw <- read.csv("~/Documents/ODSi/ODSiData/Vallet2023/Metadata/Raw_metadata/meta_samples.csv",
                     stringsAsFactors = FALSE, check.names = FALSE)

vallet_clinical_raw <- readxl::read_excel("~/Documents/ODSi/ODSiData/Vallet2023/Metadata/Raw_metadata/meta_clinical.xlsx", col_types = "text")
```

There are some corrupted dates in the xlsx file so we'll need to fix the date parsing further. Specifically, sample time should align with the azimut_id suffix and the j.cond and j.hsct should give the difference between the sample time and those 2 dates. We'll confirm that parse_mixed_dates is working correctly in this case.

```{r}
vallet_clinical_raw %>%
  mutate(
    across(
      c(sample.time, starts_with("dat")),
      ~parse_mixed_dates(.x, corrupted = T)
    )
  )%>% 
  #check sample time
  mutate(sample.time2 = dmy(str_extract(azimut_id, "\\d{2}_\\d{2}_\\d{2}$")))%>%
  #check hsct and cond
  mutate(dat.hsct2 = sample.time2 - as.numeric(j.hsct))%>%
  mutate(dat.cond2 = sample.time2 - as.numeric(j.cond))%>%
   select(azimut_id, sample.time, sample.time2,
          dat.hsct, dat.hsct2,
          dat.cond, dat.cond2)%>%
    filter((sample.time != sample.time2) |
             (dat.hsct != dat.hsct2) |
             (dat.cond != dat.cond2)
           )

#should be table with 0
#ignore parsing warnings as it is expected
```

```{r, warning=FALSE}

vallet_clinical <- vallet_clinical_raw%>%
  mutate(
    across(
      c(sample.time, starts_with("dat")),
      ~parse_mixed_dates(.x, corrupted = T)
    )
  )
```

```{r}

vallet_samples <- clean_names(vallet_samples_raw)
vallet_samples <- vallet_samples %>%
  rowwise() %>%
  mutate(sample_name_prefix = str_split(sample_name, "_")[[1]][1],
         sample_name_suffix = str_split(sample_name, "_")[[1]][2])%>%
  mutate(sample.time = parse_mixed_dates(as.numeric(sample_name_suffix)))%>%
  ungroup()

```

## Linking samples and clinical

The observations available in both data frames are age, sex, and treatment assignment. Linking these variables (age rounded to 0 decimal points), as well as the sample.time (taken directly from the clinical data frame or taken from the sample_name suffix in the samples data frame) gives us a mapping of 50 allozithro_id to sample_name_prefix, which accounts for all 50 observations in the sample.

```{r}
linking_vallet_samples <- vallet_samples %>%
  mutate(age = round(as.numeric(age), 0))%>%
  select(run, 
         sample_name_prefix,
         sample.time,
         age,
         sex,
         treatment
         )%>%
  group_by(sample_name_prefix)%>%
  mutate(no.obs = n())%>%
  ungroup()
```

```{r}
linking_vallet_clinical <- vallet_clinical %>%
  mutate(sex = ifelse(gender == 1, "male", "female"))%>%
  mutate(age = round(as.numeric(age),0 ))%>%
  select(azimut_id,
         allozithro_id,
         sample.time,
         dat.hsct,
         age,
         sex,
         treatment = group
         )%>%
  group_by(allozithro_id)%>%
  mutate(no.obs = n())%>%
  ungroup()
```

```{r}
vallet_meta <- 
```

```{r}
linking_vallet <- distinct(linking_vallet_clinical %>%
           select(allozithro_id,
                  azimut_id,
                  sample.time,
                  age,
                  sex, 
                  treatment))%>%
  full_join(linking_vallet_samples %>%
              select(run, sample_name_prefix,
                     sample.time,
                     age,
                     sex,
                     treatment),
            by = c("sample.time", "age", "sex", "treatment"))
```

```{r}
linked_vallet <- linking_vallet %>% 
  filter(!is.na(allozithro_id) & !is.na(sample_name_prefix))%>%
  select(run, sample_name_prefix, 
         azimut_id, allozithro_id, 
         sample.time) %>%
  distinct()
```

# Time invariant

```{r}
vallet_meta_qiime <- linked_vallet %>% 
  left_join(vallet_clinical,
            by = c("azimut_id", "allozithro_id", "sample.time"))%>%
  dplyr::rename(`sample-id` = run)
```

```{r}
write_tsv(vallet_meta_qiime, file = "~/Documents/ODSi/ODSiData/Vallet2023/Metadata/vallet_meta_qiime.tsv")
```
