---
title: "Bib"
format: html
editor: visual
---

# Post Query

-   Imported bibliographies from Web of science and pubmed and hand curated into papers.

    -   993 + 330 + 37

-   Made a single list of all entries with papers native deduplicate function on.

    -   1037

-   Used papers native update metadata.

-   Exported a .bib.

# Libraries

```{r}
install.packages("bib2df", "synthesisr")
library(tidyverse)
library(bib2df)
library(rcrossref)
library(synthesisr)
library(stringdist)
```

# Deduplicate

```{r}
original <- bib2df("~/Documents/ODSi/ODSiData/AI_assisted_litreview/searches/22April2026/Updated_Library.bib", separate_names = T)
```

```{r}
duplicates <- find_duplicates(original$TITLE, method = "string_osa", threshold = 5, rm_punctuation = T)
clean_refs <- extract_unique_references(original, duplicates)
bib2df::df2bib(clean_refs, file = "~/Documents/ODSi/ODSiData/AI_assisted_litreview/searches/22April2026/deduplicated.bib")
```

Now imported back into papers for a nested 839 papers.

```{r}
rm(list=ls())
```

# Cross reference

```{r}
deduplicated <- bib2df("~/Documents/ODSi/ODSiData/AI_assisted_litreview/searches/22April2026/deduplicated.bib")
```

## Get DOI

```{r}
# Robust Crossref DOI lookup for incomplete bibliographic records
# Requires: rcrossref, stringdist, dplyr, purrr
 
library(rcrossref)
library(stringdist)
library(dplyr)
library(purrr)
 
 
# ---- helpers ----------------------------------------------------------------
 
# Strip HTML/markup tags and collapse whitespace so comparisons aren't broken
# by things like "<em>Escherichia coli</em>".
clean_text <- function(x) {
  if (is.null(x) || length(x) == 0) return(NA_character_)
  x <- as.character(x)[1]
  if (is.na(x)) return(NA_character_)
  x <- gsub("<[^>]+>", " ", x)            # remove HTML tags
#  x <- gsub("[\\[\\]{}]", " ", x)         # remove brackets often added to translated titles
  x <- gsub("\\s+", " ", x)
  trimws(x)
}
 
# Pull a usable last name out of whatever AUTHOR contains (list, "Last, First",
# "First Last", "A and B and C", "A; B", etc.). Returns NA when nothing usable.
first_author_lastname <- function(authors) {
  if (is.null(authors) || length(authors) == 0) return(NA_character_)
  if (is.list(authors)) authors <- unlist(authors, use.names = FALSE)
  s <- as.character(authors)[1]
  if (is.na(s) || !nzchar(s)) return(NA_character_)
 
  # Split on common multi-author separators and take the first author
  first <- strsplit(s, "\\s+and\\s+|\\s*;\\s*|\\s*&\\s*")[[1]][1]
  first <- trimws(first)
 
  # "Last, First" -> "Last"
  if (grepl(",", first)) {
    return(trimws(strsplit(first, ",")[[1]][1]))
  }
  # "First Middle Last" -> "Last"
  parts <- strsplit(first, "\\s+")[[1]]
  if (length(parts) == 0) return(NA_character_)
  tail(parts, 1)
}
 
# Score how well a Crossref candidate matches the local record.
# Returns a value in [0, 1]; only fields that are present on BOTH sides count
# toward the weighted average, so missing data doesn't drag the score down.
score_candidate <- function(candidate,
                            title, author_last, journal, year, volume, pages) {
  w <- c(title = 0.45, author = 0.20, year = 0.15,
         journal = 0.15, volume = 0.025, pages = 0.025)
  acc <- 0; tot <- 0
 
  # title (Jaro-Winkler is robust to small edits)
  ct <- clean_text(candidate$title)
  if (!is.na(title) && nzchar(title) && !is.na(ct)) {
    sim <- stringdist::stringsim(tolower(title), tolower(ct), method = "jw")
    acc <- acc + w[["title"]] * sim; tot <- tot + w[["title"]]
  }
 
  # author last name appears anywhere in the candidate's author list
  if (!is.na(author_last) && nzchar(author_last) && !is.null(candidate$author)) {
    a <- candidate$author
    if (is.list(a)) a <- a[[1]]
    fams <- if (is.data.frame(a) && "family" %in% names(a)) a$family else a
    hit <- any(tolower(fams) == tolower(author_last), na.rm = TRUE)
    acc <- acc + w[["author"]] * as.numeric(isTRUE(hit)); tot <- tot + w[["author"]]
  }
 
  # year (exact = 1, off-by-one = 0.5, else 0 — handles online vs print year drift)
  if (!is.na(year) && nzchar(as.character(year))) {
    ty <- suppressWarnings(as.integer(year))
    cy_str <- as.character(candidate$issued %||% candidate$created %||% NA)
    cy <- suppressWarnings(as.integer(substr(cy_str, 1, 4)))
    if (!is.na(ty) && !is.na(cy)) {
      ym <- if (ty == cy) 1 else if (abs(ty - cy) == 1) 0.5 else 0
      acc <- acc + w[["year"]] * ym; tot <- tot + w[["year"]]
    }
  }
 
  # journal name
  cj <- candidate$container.title %||% candidate$`container-title`
  if (is.list(cj)) cj <- cj[[1]]
  cj <- clean_text(cj)
  if (!is.na(journal) && nzchar(journal) && !is.na(cj)) {
    sim <- stringdist::stringsim(tolower(journal), tolower(cj), method = "jw")
    acc <- acc + w[["journal"]] * sim; tot <- tot + w[["journal"]]
  }
 
  # volume
  cv <- as.character(candidate$volume %||% NA)
  if (!is.na(volume) && nzchar(as.character(volume)) && !is.na(cv) && nzchar(cv)) {
    acc <- acc + w[["volume"]] * as.numeric(as.character(volume) == cv)
    tot <- tot + w[["volume"]]
  }
 
  # first page
  cp <- as.character(candidate$page %||% NA)
  if (!is.na(pages) && nzchar(as.character(pages)) && !is.na(cp) && nzchar(cp)) {
    tp <- strsplit(as.character(pages), "[-–]")[[1]][1]
    cp1 <- strsplit(cp, "[-–]")[[1]][1]
    acc <- acc + w[["pages"]] * as.numeric(trimws(tp) == trimws(cp1))
    tot <- tot + w[["pages"]]
  }
 
  if (tot == 0) 0 else acc / tot
}
 
# tiny null-coalesce helper
`%||%` <- function(a, b) if (is.null(a) || length(a) == 0 || all(is.na(a))) b else a
 
 
# ---- main function ----------------------------------------------------------
 
fetch_and_verify_doi <- function(title = NA, authors = NA, journal = NA,
                                 year = NA, volume = NA, issue = NA,
                                 pages = NA, issn = NA,
                                 threshold = 0.80,
                                 limit = 5,
                                 verbose = TRUE) {
 
  title  <- clean_text(title)
  journal <- clean_text(journal)
  author_last <- first_author_lastname(authors)
 
  # Need *something* to search on
  if ((is.na(title) || !nzchar(title)) &&
      (is.na(author_last) || !nzchar(author_last))) {
    return(NA_character_)
  }
 
  Sys.sleep(0.3)  # be polite to Crossref
 
  # Build query: prefer field-specific queries when possible; they're far more
  # selective than free-text and dramatically reduce the "everything matches
  # Soziale Innovation" failure mode.
  args <- list(
    limit  = limit,
    select = c("title", "DOI", "author", "issued",
               "container-title", "issue", "page", "ISSN")
  )
  if (!is.na(title)        && nzchar(title))        args$query.title            <- title
  if (!is.na(author_last)  && nzchar(author_last))  args$query.author           <- author_last
  if (!is.na(journal)      && nzchar(journal))      args$query.container_title  <- journal
 
  # Year and ISSN as filters narrow the search index, not just the ranking
  filt <- character()
  yr <- suppressWarnings(as.integer(year))
  if (!is.na(yr)) {
    filt <- c(filt,
              sprintf("from-pub-date:%d", yr - 1),
              sprintf("until-pub-date:%d", yr + 1))
  }
  if (!is.na(issn) && nzchar(as.character(issn))) {
    # Crossref accepts ISSN with or without hyphen
    filt <- c(filt, sprintf("issn:%s", as.character(issn)))
  }
  if (length(filt) > 0) args$filter <- filt
 
  # Run the query (handle network errors gracefully)
  res <- tryCatch(do.call(rcrossref::cr_works, args)$data,
                  error = function(e) {
                    if (verbose) message("API error: ", conditionMessage(e))
                    NULL
                  })
  if (is.null(res) || nrow(res) == 0) return(NA_character_)
 
  # Score every candidate, pick the best
  scores <- vapply(seq_len(nrow(res)), function(i) {
    score_candidate(res[i, ], title, author_last, journal, year, volume, pages)
  }, numeric(1))
 
  best <- which.max(scores)
  best_score <- scores[best]
 
  if (is.na(best_score) || best_score < threshold) {
    if (verbose) {
      message(sprintf("No confident match (best=%.2f) for: '%s' -- top candidate: '%s'",
                      best_score,
                      substr(title %||% "", 1, 70),
                      substr(clean_text(res$title[best]) %||% "", 1, 70)))
    }
    return(NA_character_)
  }
  as.character(res$doi[best])
}
 
 
```

```{r}
fetch_and_verify_doi("A novel and effective oral antibiotics treatment to the graft-versus-host disease model mouse.")

fetch_and_verify_doi("A Novel Small Molecule Therapeutic for Acute Graft Versus Host Disease")
```

```{r}
deduplicated[c(16,17),]%>%
  # Create a logical flag for missing DOIs (bib2df usually imports them as NA if missing)
  mutate(needs_doi = is.na(DOI)) %>%
  # Map the function over the titles, but only for those missing DOIs
  mutate(
    new_doi = if_else(needs_doi, 
                      map_chr(TITLE, fetch_and_verify_doi), NA_character_),
    # Coalesce merges the original DOI column with the newly found DOIs
    DOI = coalesce(DOI, new_doi)
  ) %>% View()
```

```{r}
doi_searched <- deduplicated%>%
  # Create a logical flag for missing DOIs (bib2df usually imports them as NA if missing)
  mutate(needs_doi = is.na(DOI)) %>%
  # Map the function over the titles, but only for those missing DOIs
  mutate(
    new_doi = if_else(needs_doi, 
                      map_chr(TITLE, fetch_and_verify_doi), NA_character_),
    # Coalesce merges the original DOI column with the newly found DOIs
    DOI = coalesce(DOI, new_doi)
  ) 
```

No new dois

## Get full metadata

```{r}
generate_merged_bib <- function(df,
                                output_file = "final_merged_library.bib",
                                sleep = 0.5,
                                verbose = TRUE) {
  
  stopifnot(is.data.frame(df))
  
  if (!requireNamespace("rcrossref", quietly = TRUE)) {
    stop("Package 'rcrossref' is required. Install it with install.packages('rcrossref').")
  }
  
  # ---------- small utilities ----------
  
  scalar <- function(x, collapse = " ") {
    if (is.null(x) || length(x) == 0) return(NA_character_)
    if (is.list(x)) x <- unlist(x, use.names = FALSE)
    
    x <- as.character(x)
    x <- x[!is.na(x)]
    x <- trimws(x)
    x <- x[nzchar(x)]
    x <- x[!toupper(x) %in% c("NA", "N/A", "NULL", "NONE")]
    
    if (length(x) == 0) NA_character_ else paste(x, collapse = collapse)
  }
  
  has_value <- function(x) {
    !is.na(scalar(x))
  }
  
  get_cell <- function(row, candidates) {
    idx <- match(tolower(candidates), tolower(names(row)))
    idx <- idx[!is.na(idx)]
    
    if (length(idx) == 0) return(NA_character_)
    
    collapse <- if (any(tolower(candidates) %in% c("author", "editor"))) {
      " and "
    } else {
      " "
    }
    
    scalar(row[[idx[1]]][[1]], collapse = collapse)
  }
  
  choose_value <- function(...) {
    vals <- list(...)
    for (x in vals) {
      sx <- scalar(x)
      if (!is.na(sx)) return(sx)
    }
    NA_character_
  }
  
 clean_text_for_bibtex <- function(x) {
  x <- scalar(x)
  if (is.na(x)) return(NA_character_)
  
  x <- gsub("[\r\n\t]+", " ", x)
  
  # Decode common HTML entities that appear in Crossref/PubMed titles.
  html_entities <- c(
    "&lt;"    = "<",
    "&gt;"    = ">",
    "&amp;"   = "&",
    "&quot;"  = "\"",
    "&apos;"  = "'",
    "&#39;"   = "'",
    "&nbsp;"  = " ",
    "&ndash;" = "-",
    "&mdash;" = "-",
    "&le;"    = "<=",
    "&ge;"    = ">="
  )
  
  for (pat in names(html_entities)) {
    x <- gsub(pat, html_entities[[pat]], x, fixed = TRUE)
  }
  
  # Numeric dash variants.
  x <- gsub("&#8211;|&#x2013;", "-", x, perl = TRUE, ignore.case = TRUE)
  x <- gsub("&#8212;|&#x2014;", "-", x, perl = TRUE, ignore.case = TRUE)
  
  # Strip real HTML tags like <em>, </em>, <i>, etc.
  # This avoids accidentally deleting things like "P < 0.05".
  x <- gsub("</?[A-Za-z][A-Za-z0-9]*(\\s+[^<>]*)?>", "", x, perl = TRUE)
  
  x <- gsub("\\s+", " ", x)
  trimws(x)
}

escape_bibtex <- function(x) {
  x <- clean_text_for_bibtex(x)
  if (is.na(x)) return(NA_character_)
  
  # Escape curly braces so BibTeX does not break.
  x <- gsub("{", "\\\\{", x, fixed = TRUE)
  x <- gsub("}", "\\\\}", x, fixed = TRUE)
  
  trimws(x)
}
  
  sanitize_key <- function(x) {
    x <- scalar(x)
    if (is.na(x)) return(NA_character_)
    
    x <- gsub("\\s+", "_", x)
    x <- gsub("[^A-Za-z0-9_:+./-]", "", x)
    
    if (nzchar(x)) x else NA_character_
  }
  
  extract_year <- function(x) {
    x <- scalar(x)
    if (is.na(x)) return(NA_character_)
    
    m <- regexpr("\\b[12][0-9]{3}\\b", x, perl = TRUE)
    if (m[1] < 0) NA_character_ else regmatches(x, m)
  }
  
  make_key <- function(row = NULL, i = NA_integer_, fields = list()) {
    if (!is.null(row)) {
      key <- sanitize_key(get_cell(row, c("BIBTEXKEY", "bibtexkey")))
      if (!is.na(key)) return(key)
    }
    
    author <- choose_value(
      fields$author,
      if (!is.null(row)) get_cell(row, c("AUTHOR", "author")) else NA_character_
    )
    
    year <- choose_value(
      fields$year,
      if (!is.null(row)) get_cell(row, c("YEAR", "year")) else NA_character_
    )
    
    title <- choose_value(
      fields$title,
      if (!is.null(row)) get_cell(row, c("TITLE", "title")) else NA_character_
    )
    
    surname <- "ref"
    if (!is.na(author)) {
      first_author <- sub("\\s+and\\s+.*$", "", author, ignore.case = TRUE)
      
      if (grepl(",", first_author)) {
        surname <- sub(",.*$", "", first_author)
      } else {
        parts <- strsplit(trimws(first_author), "\\s+")[[1]]
        parts <- parts[nzchar(parts)]
        if (length(parts) > 0) surname <- parts[length(parts)]
      }
    }
    
    title_word <- ""
    if (!is.na(title)) {
      words <- strsplit(gsub("[^A-Za-z0-9]+", " ", title), "\\s+")[[1]]
      words <- words[nzchar(words)]
      if (length(words) > 0) title_word <- words[1]
    }
    
    key <- paste0(surname, if (!is.na(year)) year else "", title_word)
    key <- sanitize_key(key)
    
    if (!is.na(key)) key else paste0("ref", i)
  }
  
  
  
  build_bibtex <- function(fields, entry_type = "article", key = "ref") {
    entry_type <- scalar(entry_type)
    if (is.na(entry_type)) entry_type <- "article"
    
    entry_type <- tolower(gsub("^@", "", entry_type))
    entry_type <- gsub("[^a-z]", "", entry_type)
    if (!nzchar(entry_type)) entry_type <- "article"
    
    key <- sanitize_key(key)
    if (is.na(key)) key <- "ref"
    
    field_order <- c(
      "author", "editor", "title", "journal", "booktitle",
      "year", "month", "volume", "number", "pages",
      "publisher", "address", "doi", "pmid", "pmcid",
      "url", "abstract"
    )
    
    ordered_names <- unique(c(
      field_order[field_order %in% names(fields)],
      setdiff(names(fields), field_order)
    ))
    
    lines <- character(0)
    
    for (nm in ordered_names) {
      val <- escape_bibtex(fields[[nm]])
      if (!is.na(val)) {
        lines <- c(lines, sprintf("\t%s = {%s}", nm, val))
      }
    }
    
    if (length(lines) == 0) {
      return(sprintf("@%s{%s\n}", entry_type, key))
    }
    
    paste0(
      "@", entry_type, "{", key, ",\n",
      paste(lines, collapse = ",\n"),
      "\n}"
    )
  }
  
  add_field_if_absent <- function(bib, field, value) {
    if (!has_value(bib) || !has_value(value)) return(bib)
    
    field <- tolower(field)
    value <- escape_bibtex(value)
    
    if (grepl(paste0("(?im)^\\s*", field, "\\s*="), bib, perl = TRUE)) {
      return(bib)
    }
    
    sub(
      "\\s*,?\\s*}\\s*$",
      sprintf(",\n\t%s = {%s}\n}", field, value),
      bib,
      perl = TRUE
    )
  }
  
  normalize_pmid <- function(x) {
    x <- scalar(x)
    if (is.na(x)) return(NA_character_)
    
    m <- regexpr("[0-9]+", x, perl = TRUE)
    if (m[1] < 0) NA_character_ else regmatches(x, m)
  }
  
  normalize_pmcid <- function(x) {
    x <- scalar(x)
    if (is.na(x)) return(NA_character_)
    
    x <- toupper(gsub("\\s+", "", x))
    x <- sub("^PMCID:?", "", x)
    
    m <- regexpr("PMC[0-9]+|[0-9]+", x, perl = TRUE)
    if (m[1] < 0) return(NA_character_)
    
    y <- regmatches(x, m)
    if (grepl("^[0-9]+$", y)) y <- paste0("PMC", y)
    
    y
  }
  
  normalize_author_field <- function(x) {
  x <- clean_text_for_bibtex(x)
  if (!has_value(x)) return(NA_character_)
  
  # Already valid BibTeX author format.
  if (grepl("\\s+and\\s+", x, ignore.case = TRUE)) {
    return(x)
  }
  
  # bib2df often prints authors as:
  # "Levine, Max Z, Gregorio, Nicole E, Jewett, Michael C"
  # Convert this back to:
  # "Levine, Max Z and Gregorio, Nicole E and Jewett, Michael C"
  parts <- trimws(strsplit(x, ",", fixed = TRUE)[[1]])
  parts <- parts[nzchar(parts)]
  
  if (length(parts) >= 2 && length(parts) %% 2 == 0) {
    last_names <- parts[seq(1, length(parts), by = 2)]
    given_names <- parts[seq(2, length(parts), by = 2)]
    return(paste(paste(last_names, given_names, sep = ", "), collapse = " and "))
  }
  
  x
}

parse_bibtex <- function(bib) {
  empty <- list(
    entry_type = "article",
    key = NA_character_,
    fields = list()
  )
  
  bib <- scalar(bib, collapse = "\n")
  if (!has_value(bib) || !grepl("^\\s*@", bib)) return(empty)
  
  open_pos <- regexpr("\\{", bib, perl = TRUE)[1]
  if (open_pos < 0) return(empty)
  
  entry_type <- trimws(sub("^\\s*@", "", substr(bib, 1, open_pos - 1)))
  
  body <- substr(bib, open_pos + 1, nchar(bib))
  body <- sub("\\s*}\\s*$", "", body, perl = TRUE)
  
  first_comma <- regexpr(",", body, fixed = TRUE)[1]
  if (first_comma < 0) return(empty)
  
  key <- trimws(substr(body, 1, first_comma - 1))
  content <- substr(body, first_comma + 1, nchar(body))
  
  chars <- strsplit(content, "", fixed = TRUE)[[1]]
  n <- length(chars)
  
  if (n == 0) {
    return(list(entry_type = entry_type, key = key, fields = list()))
  }
  
  collapse_range <- function(a, b) {
    if (a > b || a > n || b < 1) return("")
    paste(chars[a:b], collapse = "")
  }
  
  fields <- list()
  pos <- 1L
  
  while (pos <= n) {
    
    # Skip whitespace and commas.
    while (
      pos <= n &&
      chars[pos] %in% c(" ", "\n", "\r", "\t", ",")
    ) {
      pos <- pos + 1L
    }
    
    if (pos > n) break
    
    # Field name.
    name_start <- pos
    
    while (
      pos <= n &&
      grepl("[A-Za-z0-9_:-]", chars[pos], perl = TRUE)
    ) {
      pos <- pos + 1L
    }
    
    if (name_start == pos) {
      pos <- pos + 1L
      next
    }
    
    name <- tolower(paste(chars[name_start:(pos - 1L)], collapse = ""))
    
    # Skip whitespace before equals.
    while (pos <= n && grepl("\\s", chars[pos], perl = TRUE)) {
      pos <- pos + 1L
    }
    
    if (pos > n || chars[pos] != "=") next
    
    pos <- pos + 1L
    
    # Skip whitespace after equals.
    while (pos <= n && grepl("\\s", chars[pos], perl = TRUE)) {
      pos <- pos + 1L
    }
    
    if (pos > n) break
    
    # Braced value: field = {...}
    if (chars[pos] == "{") {
      pos <- pos + 1L
      value_start <- pos
      depth <- 1L
      
      while (pos <= n && depth > 0) {
        ch <- chars[pos]
        
        # Ignore escaped braces while finding the matching closing brace.
        if (ch == "\\") {
          pos <- pos + 2L
          next
        }
        
        if (ch == "{") depth <- depth + 1L
        if (ch == "}") depth <- depth - 1L
        
        pos <- pos + 1L
      }
      
      value_end <- pos - 2L
      value <- collapse_range(value_start, value_end)
      
    # Quoted value: field = "..."
    } else if (chars[pos] == "\"") {
      pos <- pos + 1L
      value_start <- pos
      
      while (pos <= n) {
        ch <- chars[pos]
        
        if (ch == "\\") {
          pos <- pos + 2L
          next
        }
        
        if (ch == "\"") break
        
        pos <- pos + 1L
      }
      
      value_end <- pos - 1L
      value <- collapse_range(value_start, value_end)
      pos <- pos + 1L
      
    # Bare value: month = Feb
    } else {
      value_start <- pos
      
      while (pos <= n && chars[pos] != ",") {
        pos <- pos + 1L
      }
      
      value_end <- pos - 1L
      value <- collapse_range(value_start, value_end)
    }
    
    # If this came from a BibTeX string we already built earlier,
    # turn escaped literal braces back into real braces before rebuilding.
    value <- gsub("\\\\([{}])", "\\1", value, perl = TRUE)
    value <- clean_text_for_bibtex(value)
    
    if (has_value(value)) {
      fields[[name]] <- value
    }
  }
  
  list(
    entry_type = entry_type,
    key = key,
    fields = fields
  )
}

row_fields_for_merge <- function(row) {
  out <- list()
  
  # Fields from the original tibble that should be preserved when present.
  field_map <- c(
    "ADDRESS"       = "address",
    "ANNOTE"        = "annote",
    "BOOKTITLE"     = "booktitle",
    "CHAPTER"       = "chapter",
    "CROSSREF"      = "crossref",
    "EDITION"       = "edition",
    "EDITOR"        = "editor",
    "HOWPUBLISHED"  = "howpublished",
    "INSTITUTION"   = "institution",
    "JOURNAL"       = "journal",
    "KEY"           = "key",
    "MONTH"         = "month",
    "NOTE"          = "note",
    "NUMBER"        = "number",
    "ORGANIZATION"  = "organization",
    "PAGES"         = "pages",
    "PUBLISHER"     = "publisher",
    "SCHOOL"        = "school",
    "SERIES"        = "series",
    "TITLE"         = "title",
    "TYPE"          = "type",
    "VOLUME"        = "volume",
    "YEAR"          = "year",
    "ISSN"          = "issn",
    "DOI"           = "doi",
    "PMID"          = "pmid",
    "PMCID"         = "pmcid",
    "ABSTRACT"      = "abstract",
    "LANGUAGE"      = "language",
    "KEYWORDS"      = "keywords",
    "URL"           = "url"
  )
  
  for (col in names(field_map)) {
    val <- get_cell(row, col)
    if (has_value(val)) {
      out[[field_map[[col]]]] <- val
    }
  }
  
  author <- normalize_author_field(get_cell(row, c("AUTHOR", "author")))
  if (has_value(author)) {
    out$author <- author
  }
  
  if (has_value(out$pmid)) {
    out$pmid <- normalize_pmid(out$pmid)
  }
  
  if (has_value(out$pmcid)) {
    out$pmcid <- normalize_pmcid(out$pmcid)
  }
  
  out
}

merge_row_into_bib <- function(bib, row, i, prefer_df_author = FALSE) {
  
  parsed <- parse_bibtex(bib)
  fields <- parsed$fields
  
  row_fields <- row_fields_for_merge(row)
  
  # Prefer the original dataframe for almost everything because it is your
  # curated source. Author is handled more cautiously because Crossref/PubMed
  # often formats it better than the flat bib2df character string.
  df_preferred_fields <- names(row_fields)
  
  if (!prefer_df_author) {
    df_preferred_fields <- setdiff(df_preferred_fields, "author")
  }
  
  for (nm in names(row_fields)) {
    if (nm %in% df_preferred_fields || !has_value(fields[[nm]])) {
      fields[[nm]] <- row_fields[[nm]]
    }
  }
  
  # Always preserve your original bib2df citation key when available.
  key <- get_cell(row, c("BIBTEXKEY", "bibtexkey"))
  if (!has_value(key)) key <- parsed$key
  if (!has_value(key)) key <- make_key(row = row, i = i, fields = fields)
  
  # Preserve the original entry category when available.
  entry_type <- get_cell(row, c("CATEGORY", "ENTRYTYPE", "entrytype"))
  if (!has_value(entry_type)) entry_type <- parsed$entry_type
  if (!has_value(entry_type)) entry_type <- "article"
  
  build_bibtex(
    fields = fields,
    entry_type = entry_type,
    key = key
  )
}
  
  # ---------- DOI lookup ----------
  
  fetch_bibtex_by_doi <- function(doi) {
    doi <- scalar(doi)
    if (is.na(doi)) return(NA_character_)
    
    tryCatch({
      out <- rcrossref::cr_cn(dois = doi, format = "bibtex")
      
      if (is.null(out) || length(out) == 0 || all(is.na(out))) {
        stop("No BibTeX returned")
      }
      
      out <- paste(as.character(out[!is.na(out)]), collapse = "\n")
      out <- trimws(out)
      
      if (!nzchar(out) || !grepl("^\\s*@", out)) {
        stop("No valid BibTeX returned")
      }
      
      out
      
    }, warning = function(w) {
      # This is the important part:
      # warnings such as "agency not found" or "404 DOI Not Found"
      # are treated as failed DOI lookups.
      NA_character_
    }, error = function(e) {
      NA_character_
    })
  }
  
  # ---------- PMID lookup through PubMed MEDLINE ----------
  
  parse_medline <- function(txt) {
    out <- list()
    current_tag <- NULL
    
    for (ln in txt) {
      if (grepl("^[A-Z0-9]{2,4}\\s*-", ln)) {
        tag <- trimws(substr(ln, 1, 4))
        val <- sub("^[A-Z0-9]{2,4}\\s*-\\s*", "", ln)
        
        out[[tag]] <- c(out[[tag]], val)
        current_tag <- tag
        
      } else if (!is.null(current_tag) && grepl("^\\s+", ln)) {
        out[[current_tag]][length(out[[current_tag]])] <- paste(
          out[[current_tag]][length(out[[current_tag]])],
          trimws(ln)
        )
      }
    }
    
    out
  }
  
  first_or_na <- function(x) {
    if (is.null(x) || length(x) == 0) NA_character_ else scalar(x[1])
  }
  
  medline_doi <- function(med) {
    candidates <- c(med$AID, med$LID)
    candidates <- candidates[!is.na(candidates)]
    candidates <- candidates[grepl("\\[doi\\]", candidates, ignore.case = TRUE)]
    
    if (length(candidates) == 0) return(NA_character_)
    
    trimws(sub("\\s*\\[doi\\]\\s*$", "", candidates[1], ignore.case = TRUE))
  }
  
  fetch_bibtex_by_pmid <- function(pmid, row, i) {
    pmid <- normalize_pmid(pmid)
    if (is.na(pmid)) return(NA_character_)
    
    tryCatch({
      url <- paste0(
        "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?",
        "db=pubmed",
        "&id=", utils::URLencode(pmid, reserved = TRUE),
        "&rettype=medline",
        "&retmode=text"
      )
      
      txt <- readLines(url, warn = FALSE)
      
      if (length(txt) == 0) stop("No PubMed record returned")
      
      med <- parse_medline(txt)
      
      if (!has_value(med$PMID)) stop("No PMID found in PubMed record")
      
      authors <- if (!is.null(med$FAU)) med$FAU else med$AU
      
      fields <- list(
        author   = if (!is.null(authors)) paste(authors, collapse = " and ") else NA_character_,
        title    = first_or_na(med$TI),
        journal  = choose_value(first_or_na(med$JT), first_or_na(med$TA)),
        year     = extract_year(first_or_na(med$DP)),
        volume   = first_or_na(med$VI),
        number   = first_or_na(med$IP),
        pages    = first_or_na(med$PG),
        doi      = medline_doi(med),
        pmid     = pmid,
        pmcid    = first_or_na(med$PMC),
        abstract = if (!is.null(med$AB)) paste(med$AB, collapse = " ") else NA_character_
      )
      
      key <- make_key(row = row, i = i, fields = fields)
      
      build_bibtex(fields, entry_type = "article", key = key)
      
    }, warning = function(w) {
      NA_character_
    }, error = function(e) {
      NA_character_
    })
  }
  
  # ---------- PMCID lookup through NCBI PMC ID Converter ----------
  
  json_string <- function(raw, name) {
    m <- regexec(
      paste0('"', name, '"\\s*:\\s*"([^"]*)"'),
      raw,
      perl = TRUE
    )
    
    hit <- regmatches(raw, m)[[1]]
    
    if (length(hit) >= 2) hit[2] else NA_character_
  }
  
  fetch_bibtex_by_pmcid <- function(pmcid, row, i) {
    pmcid <- normalize_pmcid(pmcid)
    if (is.na(pmcid)) return(NA_character_)
    
    tryCatch({
      url <- paste0(
        "https://www.ncbi.nlm.nih.gov/pmc/utils/idconv/v1.0/?",
        "ids=", utils::URLencode(pmcid, reserved = TRUE),
        "&format=json"
      )
      
      raw <- paste(readLines(url, warn = FALSE), collapse = "")
      
      if (!nzchar(raw)) stop("No PMCID conversion record returned")
      
      converted_doi <- json_string(raw, "doi")
      converted_pmid <- json_string(raw, "pmid")
      
      converted_doi <- gsub("\\\\/", "/", converted_doi)
      
      # PMCID can point us to a DOI or PMID.
      # Try DOI first, then PMID.
      bib <- fetch_bibtex_by_doi(converted_doi)
      
      if (!has_value(bib)) {
        bib <- fetch_bibtex_by_pmid(converted_pmid, row = row, i = i)
      }
      
      bib
      
    }, warning = function(w) {
      NA_character_
    }, error = function(e) {
      NA_character_
    })
  }
  
  # ---------- final fallback: use current tibble row ----------
  
  row_to_bibtex <- function(row, i) {
    entry_type <- get_cell(row, c("CATEGORY", "ENTRYTYPE", "entrytype"))
    if (!has_value(entry_type)) entry_type <- "article"
    
    fields <- list()
    
    skip <- c("category", "entrytype", "bibtexkey")
    
    for (nm in names(row)) {
      if (tolower(nm) %in% skip) next
      
      val <- get_cell(row, nm)
      if (!has_value(val)) next
      
      field_name <- tolower(gsub("[^A-Za-z0-9_:-]", "", nm))
      
      # Normalize a few common non-BibTeX-ish variants
      if (field_name %in% c("pubmed", "pubmedid", "pubmed_id")) {
        field_name <- "pmid"
      }
      
      if (field_name %in% c("pmc", "pmc_id")) {
        field_name <- "pmcid"
      }
      
      fields[[field_name]] <- val
    }
    
    key <- make_key(row = row, i = i, fields = fields)
    
    build_bibtex(fields, entry_type = entry_type, key = key)
  }
  
  # ---------- main loop ----------
  
  bib_entries <- list()
  
  cat("Starting processing for", nrow(df), "entries...\n")
  
  for (i in seq_len(nrow(df))) {
    
    row <- df[i, , drop = FALSE]
    
    current_doi <- get_cell(row, c("DOI", "doi"))
    current_pmid <- get_cell(row, c("PMID", "pmid", "PUBMED", "PUBMEDID", "PUBMED_ID"))
    current_pmcid <- get_cell(row, c("PMCID", "pmcid", "PMC", "pmc", "PMC_ID"))
    current_abstract <- get_cell(row, c("ABSTRACT", "abstract", "AB"))
    
    if (verbose) {
      cat(sprintf("Processing [%s/%s]: ", i, nrow(df)))
    }
    
    bib <- NA_character_
    source_used <- "current tibble row"
    
    # 1. Try DOI first
    if (has_value(current_doi)) {
      if (verbose) cat("trying DOI... ")
      Sys.sleep(sleep)
      
      bib <- fetch_bibtex_by_doi(current_doi)
      
      if (has_value(bib)) {
        source_used <- "DOI"
      }
    }
    
    # 2. If DOI is missing, warns, errors, or returns no usable BibTeX, try PMID
    if (!has_value(bib) && has_value(current_pmid)) {
      if (verbose) cat("trying PMID... ")
      Sys.sleep(sleep)
      
      bib <- fetch_bibtex_by_pmid(current_pmid, row = row, i = i)
      
      if (has_value(bib)) {
        source_used <- "PMID"
      }
    }
    
    # 3. If PMID is missing or fails, try PMCID
    if (!has_value(bib) && has_value(current_pmcid)) {
      if (verbose) cat("trying PMCID... ")
      Sys.sleep(sleep)
      
      bib <- fetch_bibtex_by_pmcid(current_pmcid, row = row, i = i)
      
      if (has_value(bib)) {
        source_used <- "PMCID"
      }
    }
    
 # 4. If all external lookups fail, we still build from the current tibble row.
external_found <- has_value(bib)

if (!external_found) {
  source_used <- "current tibble row"
}

# Critical fix:
# Merge the original tibble row back into whatever was fetched.
# This preserves your BIBTEXKEY, title, pages, ISSN, PMID, PMCID,
# abstract, keywords, language, etc.
bib <- merge_row_into_bib(
  bib = bib,
  row = row,
  i = i,
  prefer_df_author = !external_found
)

bib_entries[[length(bib_entries) + 1]] <- bib

    if (verbose) {
      cat("used", source_used, "\n")
    }
  }
  
  valid_entries <- unlist(bib_entries, use.names = FALSE)
  
  writeLines(
    paste(valid_entries, collapse = "\n\n"),
    con = output_file,
    useBytes = TRUE
  )
  
  cat(
    "\nDone! Successfully wrote",
    length(valid_entries),
    "entries to",
    output_file,
    "\n"
  )
  
  invisible(valid_entries)
}
```

```{r}
test_set <- bind_rows(doi_searched %>% head(6), doi_searched %>% filter(is.na(ABSTRACT))%>% head())

```

```{r}
generate_merged_bib(test_set, "new_test.bib")
```

```{r}
test_read <- bib2df("~/Documents/ODSi/ODSiData/AI_assisted_litreview/searches/22April2026/new_test.bib")
```

```{r}
filled <- generate_merged_bib(doi_searched, "filled.bib")
```

```{r}
filled <- bib2df("~/Documents/ODSi/ODSiData/AI_assisted_litreview/searches/22April2026/filled.bib")
```

# Filter

Remove ref where journal is missing

```{r}
cleaned <- filled %>%
  filter(!is.na(JOURNAL))
```

```{r}
df2bib(cleaned, "~/Documents/ODSi/ODSiData/AI_assisted_litreview/searches/22April2026/cleaned.bib")
```

# After first pass

```{r}
first_pass <- readxl::read_xlsx("~/Documents/ODSi/ODSiData/AI_assisted_litreview/Screening1_AI/InitialList657_classified.xlsx")

linking_cleaned <- read_csv("~/Documents/ODSi/ODSiData/AI_assisted_litreview/Screening1_AI/InitialList657.csv")

first_pass <- linking_cleaned %>%
  left_join(first_pass,
            by = c("Title", "Abstract"))
```

```{r}
table(first_pass$Decision)
```

```{r}
table(first_pass %>% 
  filter(Decision == "Discard")%>%
  pull(Reasoning))
```

```{r}
discarded <- first_pass %>% 
  filter(Decision == "Discard")%>%
  pull(Title)
```

```{r}
# returns names of columns that are unique IDs
names(first_pass)[sapply(first_pass, function(x) length(unique(x)) == nrow(first_pass))]

```

```{r}
names(cleaned)[sapply(cleaned, function(x) length(unique(x)) == nrow(cleaned))]
```

```{r}
length(intersect(cleaned$BIBTEXKEY, first_pass$Citekey))
length(intersect(cleaned$TITLE, first_pass$Title))
```

```{r}
first_pass_included <- first_pass %>%
  filter(Decision == "Include")%>%
  select(-c(Decision, Reasoning))
```

```{r}
save(first_pass_included, file = "~/Documents/ODSi/ODSiData/AI_assisted_litreview/Screening1_AI/screened481.csv" )
```

```{r}
generate_merged_bib(first_pass_included,
                                output_file = "~/Documents/ODSi/ODSiData/AI_assisted_litreview/Screening1_AI/screened.bib",
                                sleep = 0.5,
                                verbose = TRUE)
```
